---
title: "Calculate multi-species trends by traits"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
bibliography: [references.json]
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/research-institute-for-nature-and-forest.csl
---

```{r setup, include=FALSE}
# Set up
library(knitr)
library(here)
opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(root.dir = here::here())

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)

# Packages
library(tidyverse)
library(brms)
library(posterior)
```

# Goal

Calculate model based species change indices for certain traits.

# Load data

```{r}
out_path <- file.path("data", "processed")

species_traits_final_nozero <- read_csv(
  file.path(out_path, "prop_data_per_trait_nozero.csv")
)
species_traits_final <- read_csv(file.path(out_path, "prop_data_per_trait.csv"))
```

# Model specification

We assume a multinomial distribution per period $j$ ($j = 1, 2$) for the number of grid squares $n$ per species $i$ ($i = 1, ..., S$).

$$
\{n_{1j}, ...n_{2j}\} \sim Multinom(\sum_{i=1}^S n_{ij}, \{p_{1j}, ...p_{2j}\})
$$

Where $p_{ij} = \frac{\lambda_{ij}}{\lambda_{1j} + ... + \lambda_{Sj}}$

$\{n_{1j}, ...n_{2j}\}$ can be interpreted as independent Poisson random variables

$$
n_{ij} \sim Poisson(\lambda_{ij})
$$

Such that we can specify the model

$$
\ln(\lambda_{ij}) = \beta_0 + \beta_{msci}X_{period} + ln(\Sigma_{i=1}^Sn_{ij}) + b_{0,{i = 1, ..., S}} + b_{1,{i = 1, ..., S}}X_{period}
$$

Where

-   $\beta_0$ the average log-number of grids for period 1
-   $\beta_{msci}$ the effect of period 2 compared to period 1 such that $MSCI = \exp(\beta_{msci})$
-   $X_{period}$ is 0 if period is 1980-2012 and 1 if period is 2013-2022
-   $ln(\Sigma_{i=1}^Sn_{ij})$ an offset used to control for the disproportionate search effort between the two periods
-   $b_{0,{i = 1, ..., S}}$ a random intercept for each species: deviation per species from $\beta_0$
-   $b_{1,{i = 1, ..., S}}$ a random slope for each species: deviation per species from $\beta_{msci}$

We set $corr(b_0, b_1) = 0$ since we want a separate (uncorrelated) deviation for both periods.

## Example 1: testing model structure

Example for trait 'Activity'. First for trait value 'DayActive'.

```{r}
# create cache directory if not yet available
dir.create(here("source", "source/brms_cache"), FALSE)
```

```{r}
dat <- species_traits_final %>%
  filter(trait_value == "DayActive")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 4000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing

fit <- brm(
  bf(n_grids ~ period + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  file = "source/brms_cache/fit",
  file_refit = "on_change"
)
```

Convergence ok. 

```{r}
plot(fit)
```

Model fit ok.

```{r}
pp_check(fit,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "period"
)
```

MSCI estimate:

```{r}
hypothesis(
  x = fit,
  "exp(periodp2013_2022) - 1 = 0",
  class = "b", alpha = 0.05
)
```

We also create a single model for all trait values in 'Activity'.

```{r}
dat2 <- species_traits_final %>%
  filter(trait_name == "Activity")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit2 <- brm(
  bf(n_grids ~ period * trait_value + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat2, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit2",
  file_refit = "on_change"
)
```

Convergence ok.

```{r}
plot(fit2)
```

Model fit ok.

```{r}
pp_check(fit2,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "period"
)
pp_check(fit2,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "trait_value"
)
```

Estimate MSCI 'DayActive':

```{r}
hypothesis(
  x = fit2,
  "exp(periodp2013_2022) - 1 = 0",
  class = "b", alpha = 0.05
)
```

Estimate MSCI 'DayNightActive':

```{r}
hypothesis(
  x = fit2,
  "exp(periodp2013_2022 + periodp2013_2022:trait_valueDayNightActive) - 1 = 0",
  class = "b", alpha = 0.05
)
```

Estimate MSCI 'NightActive':

```{r}
hypothesis(
  x = fit2,
  "exp(periodp2013_2022 + periodp2013_2022:trait_valueNightActive) - 1 = 0",
  class = "b", alpha = 0.05
)
```

### Conclusion

The models perform well. We have a good fit and convergence. We can use these models to calculate the MSCI.

## Example 2: Simpson's paradox

What if we compare two traits?

We look at 'Phagy' and 'EllenbergN' separately. Afterwards we combine the traits. First we look at 'Phagy'.

```{r}
## Phagy
dat_phagy <- species_traits_final %>%
  filter(trait_name == "Phagy")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit_phag <- brm(
  bf(n_grids ~ period * trait_value + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat_phagy, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit_phag",
  file_refit = "on_change"
)
```

```{r}
# Estimate MSCI 'Monophagous'
msci_monophagous <- hypothesis(
  x = fit_phag,
  "exp(periodp2013_2022) - 1 = 0",
  class = "b", alpha = 0.05
)

# Estimate MSCI 'Oligophagous'
msci_oligophagous <- hypothesis(
  x = fit_phag,
  "exp(periodp2013_2022 + periodp2013_2022:trait_valueOligophagous) - 1 = 0",
  class = "b", alpha = 0.05
)

# Estimate MSCI 'Polyphagous'
msci_polyphagous <- hypothesis(
  x = fit_phag,
  "exp(periodp2013_2022 + periodp2013_2022:trait_valuePolyphagous) - 1 = 0",
  class = "b", alpha = 0.05
)
```

```{r}
estimate_msci_monophagous <- msci_monophagous$hypothesis$Estimate
se_msci_monophagous <- msci_monophagous$hypothesis$Est.Error
ll_msci_monophagous <- msci_monophagous$hypothesis$CI.Lower
ul_msci_monophagous <- msci_monophagous$hypothesis$CI.Upper

estimate_msci_oligophagous <- msci_oligophagous$hypothesis$Estimate
se_msci_oligophagous <- msci_oligophagous$hypothesis$Est.Error
ll_msci_oligophagous <- msci_oligophagous$hypothesis$CI.Lower
ul_msci_oligophagous <- msci_oligophagous$hypothesis$CI.Upper

estimate_msci_polyphagous <- msci_polyphagous$hypothesis$Estimate
se_msci_polyphagous <- msci_polyphagous$hypothesis$Est.Error
ll_msci_polyphagous <- msci_polyphagous$hypothesis$CI.Lower
ul_msci_polyphagous <- msci_polyphagous$hypothesis$CI.Upper


# Create dataframe
phagy_summary_df <- tibble(
  est_msci = c(
    estimate_msci_monophagous,
    estimate_msci_oligophagous,
    estimate_msci_polyphagous
  ),
  se_msci = c(
    se_msci_monophagous,
    se_msci_oligophagous,
    se_msci_polyphagous
  ),
  ll_msci = c(
    ll_msci_monophagous,
    ll_msci_oligophagous,
    ll_msci_polyphagous
  ),
  ul_msci = c(
    ul_msci_monophagous,
    ul_msci_oligophagous,
    ul_msci_polyphagous
  ),
  trait_value = c("Monophagous", "Oligophagous", "Polyphagous")
)

# Plot
phagy_summary_df %>%
  ggplot(aes(y = est_msci, x = trait_value, colour = trait_value)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ll_msci, ymax = ul_msci), width = 0.1) +
  labs(x = "", y = "Multispecies Change Index", title = "Phagy") +
  theme(legend.position = "")
```

Then at 'EllenbergN':

```{r}
## EllenbergN
dat_ellenbergy <- species_traits_final %>%
  filter(trait_name == "EllenbergN")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit_ellenberg <- brm(
  bf(n_grids ~ period * trait_value + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat_ellenbergy, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit_ellenberg",
  file_refit = "on_change"
)
```

```{r}
# Estimate MSCI 'NutrientPoor'
msci_nutrientpoor <- hypothesis(
  x = fit_ellenberg,
  "exp(periodp2013_2022) - 1 = 0",
  class = "b", alpha = 0.05
)

# Estimate MSCI 'NutrientRich'
msci_nutrientrich <- hypothesis(
  x = fit_ellenberg,
  "exp(periodp2013_2022 + periodp2013_2022:trait_valueNutrientRich) - 1 = 0",
  class = "b", alpha = 0.05
)

# Estimate MSCI 'VeryNutrientPoor'
msci_verynutrientpoor <- hypothesis(
  x = fit_ellenberg,
  paste(
    "exp(periodp2013_2022 +",
    "periodp2013_2022:trait_valueVeryNutrientPoor) - 1 = 0"
  ),
  class = "b", alpha = 0.05
)

# Estimate MSCI 'VeryNutrientRich'
msci_verynutrientrich <- hypothesis(
  x = fit_ellenberg,
  paste(
    "exp(periodp2013_2022 +",
    "periodp2013_2022:trait_valueVeryNutrientRich) - 1 = 0"
  ),
  class = "b", alpha = 0.05
)

estimate_msci_nutrientpoor <- msci_nutrientpoor$hypothesis$Estimate
se_msci_nutrientpoor <- msci_nutrientpoor$hypothesis$Est.Error
ll_msci_nutrientpoor <- msci_nutrientpoor$hypothesis$CI.Lower
ul_msci_nutrientpoor <- msci_nutrientpoor$hypothesis$CI.Upper

estimate_msci_nutrientrich <- msci_nutrientrich$hypothesis$Estimate
se_msci_nutrientrich <- msci_nutrientrich$hypothesis$Est.Error
ll_msci_nutrientrich <- msci_nutrientrich$hypothesis$CI.Lower
ul_msci_nutrientrich <- msci_nutrientrich$hypothesis$CI.Upper

estimate_msci_verynutrientpoor <- msci_verynutrientpoor$hypothesis$Estimate
se_msci_verynutrientpoor <- msci_verynutrientpoor$hypothesis$Est.Error
ll_msci_verynutrientpoor <- msci_verynutrientpoor$hypothesis$CI.Lower
ul_msci_verynutrientpoor <- msci_verynutrientpoor$hypothesis$CI.Upper

estimate_msci_verynutrientrich <- msci_verynutrientrich$hypothesis$Estimate
se_msci_verynutrientrich <- msci_verynutrientrich$hypothesis$Est.Error
ll_msci_verynutrientrich <- msci_verynutrientrich$hypothesis$CI.Lower
ul_msci_verynutrientrich <- msci_verynutrientrich$hypothesis$CI.Upper


# Create dataframe
ellenberg_summary_df <- tibble(
  est_msci = c(
    estimate_msci_nutrientpoor,
    estimate_msci_nutrientrich,
    estimate_msci_verynutrientpoor,
    estimate_msci_verynutrientrich
  ),
  se_msci = c(
    se_msci_nutrientpoor,
    se_msci_nutrientrich,
    se_msci_verynutrientpoor,
    se_msci_verynutrientrich
  ),
  ll_msci = c(
    ll_msci_nutrientpoor,
    ll_msci_nutrientrich,
    ll_msci_verynutrientpoor,
    ll_msci_verynutrientrich
  ),
  ul_msci = c(
    ul_msci_nutrientpoor,
    ul_msci_nutrientrich,
    ul_msci_verynutrientpoor,
    ul_msci_verynutrientrich
  ),
  trait_value = factor(
    c(
      "NutrientPoor", "NutrientRich", "VeryNutrientPoor",
      "VeryNutrientRich"
    ),
    levels = c(
      "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
      "VeryNutrientRich"
    ), ordered = TRUE
  )
)

# Plot
ellenberg_summary_df %>%
  ggplot(aes(y = est_msci, x = trait_value, colour = trait_value)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ll_msci, ymax = ul_msci), width = 0.1) +
  labs(x = "", y = "Multispecies Change Index", title = "EllenbergN") +
  theme(legend.position = "")
```

Do the trends change when the groups are combined (= Simpson's paradox)?

```{r}
dat3 <- species_traits_final %>%
  filter(trait_name %in% c("EllenbergN", "Phagy")) %>%
  pivot_wider(
    id_cols = c(species_nl, n_grids, period, sum_per_period),
    names_from = trait_name, values_from = trait_value
  )

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit3 <- brm(
  bf(n_grids ~ period * Phagy * EllenbergN + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat3, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit3",
  file_refit = "on_change"
)
```

```{r}
msci_nutrientpoor_phagy <- hypothesis(
  x = fit3,
  c(
    "exp(periodp2013_2022) - 1 = 0",
    "exp(periodp2013_2022 + periodp2013_2022:PhagyOligophagous) - 1 = 0",
    "exp(periodp2013_2022 + periodp2013_2022:PhagyPolyphagous) - 1 = 0"
  ),
  class = "b", alpha = 0.05
)

msci_verynutrientpoor_phagy <- hypothesis(
  x = fit3,
  c(
    paste(
      "exp(periodp2013_2022 +",
      "periodp2013_2022:EllenbergNVeryNutrientPoor) - 1 = 0"
    ),
    paste(
      "exp(periodp2013_2022 +",
      "periodp2013_2022:EllenbergNVeryNutrientPoor +",
      "periodp2013_2022:PhagyOligophagous +",
      "periodp2013_2022:PhagyOligophagous:EllenbergNVeryNutrientPoor)",
      "- 1 = 0"
    ),
    paste(
      "exp(periodp2013_2022 +",
      "periodp2013_2022:EllenbergNVeryNutrientPoor +",
      "periodp2013_2022:PhagyPolyphagous +",
      "periodp2013_2022:PhagyPolyphagous:EllenbergNVeryNutrientPoor)",
      "- 1 = 0"
    )
  ),
  class = "b", alpha = 0.05
)

msci_nutrientrich_phagy <- hypothesis(
  x = fit3,
  c(
    "exp(periodp2013_2022 + periodp2013_2022:EllenbergNNutrientRich) - 1 = 0",
    paste(
      "exp(periodp2013_2022 + periodp2013_2022:EllenbergNNutrientRich +",
      "periodp2013_2022:PhagyOligophagous +",
      "periodp2013_2022:PhagyOligophagous:EllenbergNNutrientRich) - 1 = 0"
    ),
    paste(
      "exp(periodp2013_2022 + periodp2013_2022:EllenbergNNutrientRich +",
      "periodp2013_2022:PhagyPolyphagous +",
      "periodp2013_2022:PhagyPolyphagous:EllenbergNNutrientRich) - 1 = 0"
    )
  ),
  class = "b", alpha = 0.05
)

msci_verynutrientrich_phagy <- hypothesis(
  x = fit3,
  c(
    paste(
      "exp(periodp2013_2022 +",
      "periodp2013_2022:EllenbergNVeryNutrientRich) - 1 = 0"
    ),
    paste(
      "exp(periodp2013_2022 +",
      "periodp2013_2022:EllenbergNVeryNutrientRich +",
      "periodp2013_2022:PhagyOligophagous +",
      "periodp2013_2022:PhagyOligophagous:EllenbergNVeryNutrientRich)",
      "- 1 = 0"
    ),
    paste(
      "exp(periodp2013_2022 +",
      "periodp2013_2022:EllenbergNVeryNutrientRich +",
      "periodp2013_2022:PhagyPolyphagous +",
      "periodp2013_2022:PhagyPolyphagous:EllenbergNVeryNutrientRich)",
      "- 1 = 0"
    )
  ),
  class = "b", alpha = 0.05
)

est_msci_nutpoor_phagy <- msci_nutpoor_phagy$hypothesis$Estimate
se_msci_nutpoor_phagy <- msci_nutpoor_phagy$hypothesis$Est.Error
ll_msci_nutpoor_phagy <- msci_nutpoor_phagy$hypothesis$CI.Lower
ul_msci_nutpoor_phagy <- msci_nutpoor_phagy$hypothesis$CI.Upper

est_msci_nutrich_phagy <- msci_nutrich_phagy$hypothesis$est
se_msci_nutrich_phagy <- msci_nutrich_phagy$hypothesis$Est.Error
ll_msci_nutrich_phagy <- msci_nutrich_phagy$hypothesis$CI.Lower
ul_msci_nutrich_phagy <- msci_nutrich_phagy$hypothesis$CI.Upper

est_msci_verynutpoor_phagy <- msci_verynutpoor_phagy$hypothesis$est
se_msci_verynutpoor_phagy <- msci_verynutpoor_phagy$hypothesis$Est.Error
ll_msci_verynutpoor_phagy <- msci_verynutpoor_phagy$hypothesis$CI.Lower
ul_msci_verynutpoor_phagy <- msci_verynutpoor_phagy$hypothesis$CI.Upper

est_msci_verynutrich_phagy <- msci_verynutrich_phagy$hypothesis$est
se_msci_verynutrich_phagy <- msci_verynutrich_phagy$hypothesis$Est.Error
ll_msci_verynutrich_phagy <- msci_verynutrich_phagy$hypothesis$CI.Lower
ul_msci_verynutrich_phagy <- msci_verynutrich_phagy$hypothesis$CI.Upper
```

```{r}
# Create dataframe
ellenberg_phagy_summary_df <- tibble(
  est_msci = c(
    est_msci_nutpoor_phagy,
    est_msci_nutrich_phagy,
    est_msci_verynutpoor_phagy,
    est_msci_verynutrich_phagy
  ),
  se_msci = c(
    se_msci_nutpoor_phagy,
    se_msci_nutrich_phagy,
    se_msci_verynutpoor_phagy,
    se_msci_verynutrich_phagy
  ),
  ll_msci = c(
    ll_msci_nutpoor_phagy,
    ll_msci_nutrich_phagy,
    ll_msci_verynutpoor_phagy,
    ll_msci_verynutrich_phagy
  ),
  ul_msci = c(
    ul_msci_nutpoor_phagy,
    ul_msci_nutrich_phagy,
    ul_msci_verynutpoor_phagy,
    ul_msci_verynutrich_phagy
  ),
  Phagy = rep(c("Monophagous", "Oligophagous", "Polyphagous"), 4),
  EllenbergN = factor(
    rep(c(
      "NutrientPoor", "NutrientRich", "VeryNutrientPoor",
      "VeryNutrientRich"
    ), each = 3),
    levels = c(
      "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
      "VeryNutrientRich"
    ), ordered = TRUE
  )
) %>%
  arrange(Phagy, EllenbergN)

ellenberg_phagy_summary_df %>%
  kable(digits = 3)
```

```{r}
# Plot
ellenberg_phagy_summary_df %>%
  ggplot(aes(y = est_msci, x = EllenbergN, colour = EllenbergN)) +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  geom_point(size = 4) +
  geom_errorbar(aes(ymin = ll_msci, ymax = ul_msci), width = 0.1) +
  labs(x = "", y = "Multispecies Change Index") +
  theme(
    legend.position = "",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  ) +
  facet_wrap(~Phagy, scales = "free_x")
```

### Conclusion

We see that monophagous species perform better in the second compared to the reference period. For oligophagous and polyphagous species there is no significant difference.
We see an increase in MSCI from poorer to richer nutrients.

When we combine both traits we see a more complicated picture. Monophagous species only perform better if they are linked to (very) nutrient rich habitats. Monophagous species of very nutrient poor habitats even have a significant negative MSCI. For oligophagous and polyphagous species we see no significant differences for any nutrient group. Only polyphagous species in nutrient poor habitats are less prevalent than in the reference period.

This phenomenon is called Simpson's paradox. We realize now that an increase in MSCI from poor to rich nutrients is only valid for monophagous species, not for oligophagous or polyphagous species. We also observe that monophagous only have a significant higher MSCI for (very) nutrient rich habitats.

To avoid erroneous conclusions related to this paradox, we will select the $t$ most important traits that we are interested in and order these from most important to least important ($1, ..., t$). We create a model with the most important trait. Afterwards we add the second trait etc. We only keep groups containing more than 10 species.
