---
title: "Calculate multi-species trends by traits"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
bibliography: [references.json]
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/research-institute-for-nature-and-forest.csl
---

```{r setup, include=FALSE}
# Set up
library(knitr)
library(here)
opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(root.dir = here::here())

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)
conflicted::conflicts_prefer(stats::lag)
conflicted::conflicts_prefer(brms::ar)
conflicted::conflicts_prefer(brms::dstudent_t)
conflicted::conflicts_prefer(brms::pstudent_t)
conflicted::conflicts_prefer(brms::qstudent_t)
conflicted::conflicts_prefer(brms::rstudent_t)

# Packages
library(tidyverse)
library(brms)
library(tidybayes)

# Source
source(here("source", "summarise_brms_models.R"))
```

# Goal

Calculate model based species change indices for certain traits.

# Load data

```{r}
data_path <- file.path("data", "processed")

species_traits_final_nozero <- read_csv(
  file.path(data_path, "prop_data_per_trait_nozero.csv")
)
species_traits_final <- read_csv(file.path(data_path,
                                           "prop_data_per_trait.csv"))
```

# Model specification

We assume a multinomial distribution per period $j$ ($j = 1, 2$) for the number of grid squares $n$ per species $i$ ($i = 1, ..., S$).

$$
\{n_{1j}, ...n_{2j}\} \sim Multinom(\sum_{i=1}^S n_{ij}, \{p_{1j}, ...p_{2j}\})
$$

Where $p_{ij} = \frac{\lambda_{ij}}{\lambda_{1j} + ... + \lambda_{Sj}}$

$\{n_{1j}, ...n_{2j}\}$ can be interpreted as independent Poisson random variables

$$
n_{ij} \sim Poisson(\lambda_{ij})
$$

Such that we can specify the model

$$
\ln(\lambda_{ij}) = \beta_0 + \beta_{msci}X_{period} + ln(\Sigma_{i=1}^Sn_{ij}) + b_{0,{i = 1, ..., S}} + b_{1,{i = 1, ..., S}}X_{period}
$$

Where

-   $\beta_0$ the average log-number of grids for period 1
-   $\beta_{msci}$ the effect of period 2 compared to period 1 such that $MSCI = \exp(\beta_{msci})$
-   $X_{period}$ is 0 if period is 1980-2012 and 1 if period is 2013-2022
-   $ln(\Sigma_{i=1}^Sn_{ij})$ an offset used to control for the disproportionate search effort between the two periods
-   $b_{0,{i = 1, ..., S}}$ a random intercept for each species: deviation per species from $\beta_0$
-   $b_{1,{i = 1, ..., S}}$ a random slope for each species: deviation per species from $\beta_{msci}$

We set $corr(b_0, b_1) = 0$ since we want a separate (uncorrelated) deviation for both periods.

## Testing model structure

Example for trait 'Activity'. First for trait value 'DayActive'.

```{r}
# create cache directory if not yet available
dir.create(here("source", "source/brms_cache"), FALSE)
```

```{r}
dat <- species_traits_final %>%
  filter(trait_value == "DayActive")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 4000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing

fit <- brm(
  bf(n_grids ~ period + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  file = "source/brms_cache/fit",
  file_refit = "on_change"
)
```

Convergence ok. 

```{r}
plot(fit)
```

Model fit ok.

```{r}
pp_check(fit,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "period"
)
```

Results:

```{r}
msci_draws_df1 <- fit %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  mutate(DayActive = exp(b_periodp2013_2022) - 1) %>%
  select(DayActive) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci")

summ_df1 <- msci_draws_df1 %>%
  group_by(trait) %>%
  mean_qi() %>%
  rename(mean = msci)

summ_df1 %>%
  kable()
```

```{r}
msci_draws_df1 %>%
  full_join(summ_df1, by = join_by(trait)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  ggplot(aes(y = msci, x = trait, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(legend.position = "")
```

We also create a single model for all trait values in 'Activity'.

```{r}
dat2 <- species_traits_final %>%
  filter(trait_name == "Activity")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit2 <- brm(
  bf(n_grids ~ period * trait_value + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat2, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit2",
  file_refit = "on_change"
)
```

Convergence ok.

```{r}
plot(fit2)
```

Model fit ok.

```{r}
pp_check(fit2,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "period"
)
pp_check(fit2,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "trait_value"
)
```

Results:

```{r}
draw_matrix <- fit2 %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

contrast_mat <- create_contrast_matrix(fit2)

msci_draws_df2 <- as_tibble(exp(t(contrast_mat %*% t(draw_matrix))) - 1) %>%
  `colnames<-`(get_traits(fit2)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci")

summ_df2 <- msci_draws_df2 %>%
  group_by(trait) %>%
  mean_qi() %>%
  rename(mean = msci)
```

```{r}
summ_df2 %>%
  rename(msci = mean) %>%
  kable()
```

```{r}
msci_draws_df2 %>%
  full_join(summ_df2, by = join_by(trait)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  ggplot(aes(y = msci, x = trait, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(legend.position = "")
```

### Conclusion

The models perform well. We have a good fit and convergence. We can use these models to calculate the MSCI.

## Exploration Simpson's paradox

What if we compare two traits? We look at 'Phagy' and 'EllenbergN'.

```{r}
species_traits_final %>%
  filter(trait_name %in% c("EllenbergN", "Phagy")) %>%
  pivot_wider(
    id_cols = c(species_nl, n_grids, period, sum_per_period),
    names_from = trait_name, values_from = trait_value
  ) %>%
  group_by(EllenbergN, Phagy) %>%
  mutate(n = n_distinct(species_nl)) %>%
  ungroup() %>%
  mutate(EllenbergN = factor(EllenbergN, levels = c(
    "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
    "VeryNutrientRich"
    ), ordered = TRUE)
  ) %>%
  ggplot(aes(x = EllenbergN, y = Phagy)) +
    geom_tile(aes(fill = n), colour = "white") +
    geom_text(aes(label = n)) +
    scale_fill_gradient(low = "blue", high = "red") +
    labs(fill = "number of\nspecies", x = "", y = "")
```

We look at 'Phagy' and 'EllenbergN' separately. Afterwards we combine the traits. First we look at 'Phagy'.

```{r}
## Phagy
dat_phagy <- species_traits_final %>%
  filter(trait_name == "Phagy")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit_phag <- brm(
  bf(n_grids ~ period * trait_value + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat_phagy, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit_phag",
  file_refit = "on_change"
)
```

Results:

```{r}
draw_matrix <- fit_phag %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

contrast_mat <- create_contrast_matrix(fit_phag)

msci_draws_df_phag <- as_tibble(exp(t(contrast_mat %*% t(draw_matrix))) - 1) %>%
  `colnames<-`(get_traits(fit_phag)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci")

summ_df_phag <- msci_draws_df_phag %>%
  group_by(trait) %>%
  mean_qi() %>%
  rename(mean = msci)
```

```{r}
summ_df_phag %>%
  rename(msci = mean) %>%
  kable()
```

```{r}
msci_draws_df_phag %>%
  full_join(summ_df_phag, by = join_by(trait)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  ggplot(aes(y = msci, x = trait, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(legend.position = "")
```

Then at 'EllenbergN':

```{r}
## EllenbergN
dat_ellenbergy <- species_traits_final %>%
  filter(trait_name == "EllenbergN")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit_ellenberg <- brm(
  bf(n_grids ~ period * trait_value + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat_ellenbergy, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit_ellenberg",
  file_refit = "on_change"
)
```

Results:

```{r}
draw_matrix <- fit_ellenberg %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

contrast_mat <- create_contrast_matrix(fit_ellenberg)

msci_draws_df_ellenberg <- as_tibble(
  exp(t(contrast_mat %*% t(draw_matrix))) - 1
  ) %>%
  `colnames<-`(get_traits(fit_ellenberg)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci")

summ_df_ellenberg <- msci_draws_df_ellenberg %>%
  group_by(trait) %>%
  mean_qi() %>%
  rename(mean = msci)
```

```{r}
summ_df_ellenberg %>%
  rename(msci = mean) %>%
  kable()
```

```{r}
msci_draws_df_ellenberg %>%
  full_join(summ_df_ellenberg, by = join_by(trait)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  mutate(trait = factor(trait, levels = c(
    "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
    "VeryNutrientRich"
  ), ordered = TRUE)) %>%
  ggplot(aes(y = msci, x = trait, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(legend.position = "")
```

Do the trends change when the groups are combined (= Simpson's paradox)?

```{r}
dat3 <- species_traits_final %>%
  filter(trait_name %in% c("EllenbergN", "Phagy")) %>%
  pivot_wider(
    id_cols = c(species_nl, n_grids, period, sum_per_period),
    names_from = trait_name, values_from = trait_value
  )

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit3 <- brm(
  bf(n_grids ~ period * Phagy * EllenbergN + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat3, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit3",
  file_refit = "on_change"
)
```

Results:

```{r}
draw_matrix <- fit3 %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

contrast_mat <- rbind(
  cbind(create_contrast_matrix(3), matrix(rep(0, 9 * 3), 3)),
  matrix(cbind(
    create_contrast_matrix(3),
    matrix(c(rep(1, 3), rep(0, 6)), ncol = 3),
    contr.treatment(3),
    matrix(rep(0, 4 * 3), 3)
  ), nrow = 3),
  matrix(cbind(
    create_contrast_matrix(3),
    matrix(c(rep(0, 3), rep(1, 3), rep(0, 3)), ncol = 3),
    matrix(rep(0, 2 * 3), 3),
    contr.treatment(3),
    matrix(rep(0, 2 * 3), 3)
  ), nrow = 3),
  matrix(cbind(
    create_contrast_matrix(3),
    matrix(c(rep(0, 6), rep(1, 3)), ncol = 3),
    matrix(rep(0, 4 * 3), 3),
    contr.treatment(3)
  ), nrow = 3)
)

ellenbergn <- sort(unique(fit3$data$EllenbergN))
phagy <- sort(unique(fit3$data$Phagy))
col_names <- lapply(ellenbergn, function(i) {
  paste(rep(i, 3), phagy, sep = "_")
})

msci_draws_ellenberg_phagy <- as_tibble(
  exp(t(contrast_mat %*% t(draw_matrix))) - 1
  ) %>%
  `colnames<-`(unlist(col_names)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci") %>%
  separate(trait, into = c("EllenbergN", "Phagy")) %>%
  mutate(EllenbergN = factor(
    EllenbergN,
    levels = c(
      "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
      "VeryNutrientRich"
    ), ordered = TRUE
  ))

summ_ellenberg_phagy <- msci_draws_ellenberg_phagy %>%
  group_by(EllenbergN, Phagy) %>%
  mean_qi() %>%
  rename(mean = msci)
```

```{r}
summ_ellenberg_phagy %>%
  select(Phagy, EllenbergN, everything()) %>%
  arrange(Phagy, EllenbergN) %>%
  rename(msci = mean) %>%
  kable()
```

```{r}
msci_draws_ellenberg_phagy %>%
  full_join(summ_ellenberg_phagy, by = join_by(EllenbergN, Phagy)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  ggplot(aes(y = msci, x = EllenbergN, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(
    legend.position = "",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  facet_wrap(~Phagy, scales = "free")
```

### Conclusion

We see that monophagous species perform better in the second compared to the reference period. For oligophagous and polyphagous species there is no significant difference.
We see an increase in MSCI from poorer to richer nutrients.

When we combine both traits we see a more complicated picture. Monophagous species only perform better if they are linked to (very) nutrient rich habitats. Monophagous species of very nutrient poor habitats even have a significant negative MSCI. For oligophagous and polyphagous species we see no significant differences for any nutrient group. Only polyphagous species in nutrient poor habitats are less prevalent than in the reference period.

This phenomenon is called Simpson's paradox. We realize now that an increase in MSCI from poor to rich nutrients is especially pronounced for monophagous species, less for oligophagous or polyphagous species. We also observe that monophagous only have a significant higher MSCI for (very) nutrient rich habitats.

To avoid erroneous conclusions related to this paradox, we will select the $t$ most important traits that we are interested in and order these from most important to least important ($1, ..., t$). We create a model with the most important trait. Afterwards we add the second trait etc. We only keep groups containing more than 10 species.

## Exclusion of zeroes

Do we see large differences when we exclude zeroes?

```{r}
dat_nozero <- species_traits_final_nozero %>%
  filter(trait_name %in% c("EllenbergN", "Phagy")) %>%
  pivot_wider(
    id_cols = c(species_nl, n_grids, period, sum_per_period),
    names_from = trait_name, values_from = trait_value
  )

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit_nozero <- brm(
  bf(n_grids ~ period * Phagy * EllenbergN + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat_nozero, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit_nozero",
  file_refit = "on_change"
)
```

Results:

```{r}
draw_matrix_nozero <- fit_nozero %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

msci_draws_nozero <- as_tibble(
  exp(t(contrast_mat %*% t(draw_matrix_nozero))) - 1
  ) %>%
  `colnames<-`(unlist(col_names)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci") %>%
  separate(trait, into = c("EllenbergN", "Phagy")) %>%
  mutate(EllenbergN = factor(
    EllenbergN,
    levels = c(
      "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
      "VeryNutrientRich"
    ), ordered = TRUE
  ))

summ_ellenberg_phagy_nozero <- msci_draws_nozero %>%
  group_by(EllenbergN, Phagy) %>%
  mean_qi() %>%
  rename(mean = msci)
```

```{r}
summ_ellenberg_phagy_nozero %>%
  select(Phagy, EllenbergN, everything()) %>%
  arrange(Phagy, EllenbergN) %>%
  rename(msci = mean) %>%
  kable()
```

```{r}
msci_draws_nozero %>%
  full_join(summ_ellenberg_phagy_nozero, by = join_by(EllenbergN, Phagy)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  ggplot(aes(y = msci, x = EllenbergN, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(
    legend.position = "",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  facet_wrap(~Phagy, scales = "free")
```

There are differences, but they are not large.
