---
title: "Calculate MSCI for resident species"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Set up
library(knitr)
library(here)
opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(root.dir = here::here())

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)
conflicted::conflicts_prefer(stats::lag)
conflicted::conflicts_prefer(brms::ar)
conflicted::conflicts_prefer(brms::dstudent_t)
conflicted::conflicts_prefer(brms::pstudent_t)
conflicted::conflicts_prefer(brms::qstudent_t)
conflicted::conflicts_prefer(brms::rstudent_t)

# Packages
library(tidyverse)
library(brms)
library(tidybayes)

# Source
source(here("source", "summarise_brms_models.R"))
source(here("source", "set_priors_brms.R"))

# Seed
seed <- 123
```

# Goal

Fit models and calculate MSCI for all resident species.

# Load data

```{r load-data}
data_path <- file.path("data", "processed")

species_traits_final <- read_csv(file.path(data_path,
                                           "prop_data_per_trait.csv"))
```

# Data preparation

```{r data-prep}
species_df_raw <- species_traits_final %>%
  filter(trait_value == "Resident") %>%
  select(species_nl, n_grids, period, sum_per_period) %>%
  mutate(species_nl = factor(species_nl))
```

In total we have `r length(unique(species_df_raw$species_nl))` resident species.
The following species have disapeared or emerged.
We cannot calculate a species change index.

```{r}
species_df_raw %>%
  pivot_wider(id_cols = species_nl,
              names_from = period,
              values_from = n_grids) %>%
  filter(p1980_2012 == 0 | p2013_2022 == 0) %>%
  kable()
```

```{r}
inf_species <- species_df_raw %>%
  pivot_wider(id_cols = species_nl,
              names_from = period,
              values_from = n_grids) %>%
  filter(p1980_2012 == 0 | p2013_2022 == 0) %>%
  pull(species_nl) %>%
  as.vector()

species_df <- species_df_raw %>%
  filter(!species_nl %in% inf_species) %>%
  mutate(species_nl = factor(species_nl))
```


# Fit model

```{r cache-dir}
# create cache directory if not yet available
dir.create(here("source", "brms_cache"), FALSE)
```

```{r mcmc-params}
# MCMC parameters
nchains <- 3 # number of chains
niter <- 20000 # number of iterations (incl. burn-in)
burnin <- niter / 5 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 8
```

```{r fit-model}
fit_msci_species <- brm(
  bf(n_grids ~ period * species_nl + offset(log(sum_per_period))),
  data = species_df, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit_msci_species",
  file_refit = "on_change"
)
```


# Model convergence

We check MCMC convergence based on posterior densities, trace plots and the potential scale reduction factors ($\hat{R}$) for each parameter in each model.

```{r trace-plots}
plot(fit_msci_species, ask = FALSE)
```

```{r rhat}
rhat_list <- get_rhats(list(species_model = fit_msci_species))
show_rhat(do.call(rbind.data.frame, rhat_list))
```

Convergence is good.

# Checking model fit

We check model fit based on posterior predictive checks where we compare the observed outcome to simulated datasets from the posterior predictive distribution.

```{r}
pp_check(fit_msci_species,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "period"
)
```

Model fit is good.


# Results

```{r summarise-draws}
# get posterior draws for parameters of interest
draws_matrix <- fit_msci_species %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE, seed = seed) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

# get contrast matrix
species <- sort(unique(species_df$species_nl))
contrast_matrix <- create_contrast_matrix(length(species))

# calculate msci
sci_df <- as_tibble(
  exp(t(contrast_matrix %*% t(draws_matrix))) - 1
  ) %>%
  `colnames<-`(species) %>%
  pivot_longer(cols = everything(), names_to = "species_nl",
               values_to = "sci")

# get summary statistics
summary_draw <- sci_df %>%
  group_by(species_nl) %>%
  mean_qi(.width = 0.95) %>%
  ungroup() %>%
  rename(sci.mean = sci,
         sci.lower = .lower,
         sci.upper = .upper)
```

```{r}
library(INBOtheme)
theme_set(theme_inbo(base_size = 12, transparent = TRUE))

min_threshold <- -0.2
max_threshold <- 0.25
limits <- c(-1.5, 10)

summary_draw %>%
  mutate(
    effects = case_when(
        sci.lower > 0 ~ "increase",
        sci.upper < 0 ~ "decrease",
        sci.lower < 0 & sci.lower > -0.2 &
          sci.upper > 0 & sci.upper < 0.25 ~ "stable",
        TRUE ~ "uncertain",
    ),
    fill_var = factor(effects,
      levels = c("increase",
                 "stable",
                 "decrease",
                 "uncertain"),
      ordered = TRUE)
  ) %>%
  arrange(fill_var, desc(sci.mean)) %>%
  rownames_to_column(var = "id") %>%
  mutate(
    id = as.numeric(id),
    species_nl = reorder(species_nl, id,  decreasing = TRUE)
    ) %>%
  ggplot(aes(y = species_nl, colour = fill_var)) +
    geom_segment(aes(x = sci.lower, xend = ifelse(sci.upper > limits[2],
                                                  limits[2], sci.upper), 
                     yend = species_nl)) +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
                 linetype = "dotdash") +
    scale_colour_manual(values =  c("chartreuse3",
                                    "gold",
                                    "firebrick1",
                                    "skyblue"),
                        drop = FALSE) +
    labs(x = "Species Change Index",
         y = "",
         colour = "Legend") +
    scale_x_continuous(limits = limits, expand = c(0, 0),
                       breaks = c(-1, 1, seq(0, 10, 2.5))) +
    theme(panel.grid.major.y = element_blank())
```

