---
title: "Calculate SCI for resident species"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Set up
library(knitr)
library(here)
opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(root.dir = here::here())

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)
conflicted::conflicts_prefer(stats::lag)
conflicted::conflicts_prefer(brms::ar)
conflicted::conflicts_prefer(brms::dstudent_t)
conflicted::conflicts_prefer(brms::pstudent_t)
conflicted::conflicts_prefer(brms::qstudent_t)
conflicted::conflicts_prefer(brms::rstudent_t)

# Packages
library(tidyverse)
library(brms)
library(tidybayes)
library(effectclass)

# Source
source(here("source", "summarise_brms_models.R"))
source(here("source", "set_priors_brms.R"))

# Seed
seed <- 123
```

# Goal

Fit models and calculate SCI for all resident species.

# Load data

```{r load-data}
data_path <- file.path("data", "processed")

species_traits_final <- read_csv(file.path(data_path,
                                           "prop_data_per_trait.csv"))
```

# Data preparation

```{r data-prep}
species_df_raw <- species_traits_final %>%
  filter(trait_value == "Resident") %>%
  select(species_nl, n_grids, period, sum_per_period) %>%
  mutate(species_nl = factor(species_nl))
```

In total we have `r length(unique(species_df_raw$species_nl))` resident species.
The following species have disappeared or emerged.
We cannot calculate a species change index for these.

```{r}
species_df_raw %>%
  pivot_wider(id_cols = species_nl,
              names_from = period,
              values_from = n_grids) %>%
  filter(p1980_2012 == 0 | p2013_2022 == 0) %>%
  kable()
```

```{r}
inf_species <- species_df_raw %>%
  pivot_wider(id_cols = species_nl,
              names_from = period,
              values_from = n_grids) %>%
  filter(p1980_2012 == 0 | p2013_2022 == 0) %>%
  pull(species_nl) %>%
  as.vector()

species_df <- species_df_raw %>%
  filter(!species_nl %in% inf_species) %>%
  mutate(species_nl = factor(species_nl))
```


# Fit model

```{r cache-dir}
# create cache directory if not yet available
dir.create(here("source", "brms_cache"), FALSE)
```

```{r mcmc-params}
# MCMC parameters
nchains <- 3 # number of chains
niter <- 20000 # number of iterations (incl. burn-in)
burnin <- niter / 5 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 8
```

```{r fit-model}
fit_msci_species <- brm(
  bf(n_grids ~ period * species_nl + offset(log(sum_per_period))),
  data = species_df, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit_msci_species",
  file_refit = "on_change"
)
```


# Model convergence

We check MCMC convergence based on posterior densities, trace plots and the potential scale reduction factors ($\hat{R}$) for each parameter in each model.

```{r trace-plots}
plot(fit_msci_species, ask = FALSE)
```

```{r rhat}
rhat_list <- get_rhats(list(species_model = fit_msci_species))
show_rhat(do.call(rbind.data.frame, rhat_list))
```

Convergence is good.

# Checking model fit

We check model fit based on posterior predictive checks where we compare the observed outcome to simulated datasets from the posterior predictive distribution.

```{r}
pp_check(fit_msci_species,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "period"
)
```

Model fit is good.


# Results

```{r summarise-draws}
# get posterior draws for parameters of interest
draws_matrix <- fit_msci_species %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE, seed = seed) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

# get contrast matrix
species <- sort(unique(species_df$species_nl))
contrast_matrix <- create_contrast_matrix(length(species))

# calculate msci
sci_df <- as_tibble(
  exp(t(contrast_matrix %*% t(draws_matrix))) - 1
  ) %>%
  `colnames<-`(species) %>%
  pivot_longer(cols = everything(), names_to = "species_nl",
               values_to = "sci")

# get summary statistics
min_threshold <- -0.2
max_threshold <- 0.25

summary_draw_pre <- sci_df %>%
  group_by(species_nl) %>%
  median_qi(.width = 0.9) %>%
  ungroup() %>%
  rename(sci.median = sci,
         sci.lower = .lower,
         sci.upper = .upper)

summary_draw_pre$effect_fine <- classification(
    lcl = summary_draw_pre$sci.lower,
    ucl = summary_draw_pre$sci.upper,
    threshold = c(min_threshold, max_threshold))

summary_draw_pre$effect_coarse <- coarse_classification(
  summary_draw$effect_fine
  )

summary_draw <- summary_draw_pre %>%
  mutate(
    trend_fine = case_when(
        effect_fine == "++" ~ "strong increase",
        effect_fine == "+" ~ "increase",
        effect_fine == "+~" ~ "moderate increase",
        effect_fine == "~" ~ "stable",
        effect_fine == "-" ~ "decrease",
        effect_fine == "-~" ~ "moderate decrease",
        effect_fine == "--" ~ "strong decrease",
        TRUE ~ "uncertain",
    ),
    trend_fine = factor(trend_fine,
      levels = c(
        "strong increase",
        "increase",
        "moderate increase",
        "stable",
        "moderate decrease",
        "decrease",
        "strong decrease",
        "uncertain"),
      ordered = TRUE)
  ) %>%
  mutate(
    trend_coarse = case_when(
        effect_coarse == "+" ~ "increase",
        effect_coarse == "-" ~ "decrease",
        effect_coarse == "~" ~ "stable",
        TRUE ~ "uncertain",
    ),
    trend_coarse = factor(trend_coarse,
      levels = c("increase",
                 "stable",
                 "decrease",
                 "uncertain"),
      ordered = TRUE)
  ) %>%
  mutate(certainty = ifelse(trend_coarse == "uncertain", "uncertain trend",
                            "certain trend"))
```

```{r}
library(INBOtheme)
theme_set(theme_inbo(base_size = 12, transparent = TRUE))

limits <- c(-1.5, 10)
```

We show the posterior median and 90 % credible intervals for each species.
For some species, the credible intervals are very large and are cut off.
Ordered according to posterior median:

```{r fig.height=10}
summary_draw %>%
  arrange(desc(sci.median)) %>%
  rownames_to_column(var = "id") %>%
  mutate(
    id = as.numeric(id),
    species_nl = reorder(species_nl, id,  decreasing = TRUE)
  ) %>%
  ggplot(aes(y = species_nl)) +
    geom_segment(aes(x = sci.lower,
                     xend = ifelse(sci.upper > limits[2],
                                   limits[2], sci.upper),
                     yend = species_nl,
                     colour = trend_fine)) +
    geom_point(aes(x = sci.median), size = 0.1, colour = "black") +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
                 linetype = "dotdash") +
    scale_colour_manual(values =  c(rev(traffic_palette(7)), "skyblue"),
                        drop = FALSE) +
    labs(x = "Species Change Index",
         y = "",
         colour = "Legend") +
    scale_x_continuous(limits = limits, expand = c(0, 0),
                       breaks = c(-1, 1, seq(0, 10, 2.5))) +
    theme(axis.text.y = element_text(size = 2),
          axis.ticks.y = element_blank()) +
    guides(colour = guide_legend(override.aes = list(linewidth = 3))) +
    facet_wrap(~certainty, scales = "free_y")
```

Ordered according to effect:

```{r fig.height=10}
summary_draw %>%
  arrange(trend_fine, desc(sci.median)) %>%
  rownames_to_column(var = "id") %>%
  mutate(
    id = as.numeric(id),
    species_nl = reorder(species_nl, id,  decreasing = TRUE)
  ) %>%
  ggplot(aes(y = species_nl)) +
    geom_segment(aes(x = sci.lower,
                     xend = ifelse(sci.upper > limits[2],
                                   limits[2], sci.upper),
                     yend = species_nl,
                     colour = trend_fine)) +
    geom_point(aes(x = sci.median), size = 0.1, colour = "black") +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
                 linetype = "dotdash") +
    scale_colour_manual(values =  c(rev(traffic_palette(7)), "skyblue"),
                        drop = FALSE) +
    labs(x = "Species Change Index",
         y = "",
         colour = "Legend") +
    scale_x_continuous(limits = limits, expand = c(0, 0),
                       breaks = c(-1, 1, seq(0, 10, 2.5))) +
    theme(axis.text.y = element_text(size = 2),
          axis.ticks.y = element_blank()) +
    guides(colour = guide_legend(override.aes = list(linewidth = 3)))
```

For the bar plots we also include the species with zero counts (but with at least 5 occurrences in the other period).
They are included in the categories 'strong increase/decrease'.

```{r}
extra_spec_df <- species_df_raw %>%
  pivot_wider(id_cols = species_nl,
              names_from = period,
              values_from = n_grids) %>%
  filter(p1980_2012 == 0 | p2013_2022 == 0) %>%
  mutate(trend_coarse = ifelse(p1980_2012 == 0,
                                 "increase",
                                 "decrease"),
         trend_fine = ifelse(p1980_2012 == 0,
                                "strong increase",
                                "strong decrease")) %>%
  mutate(
    trend_coarse = factor(trend_coarse,
      levels = c("increase",
                 "stable",
                 "decrease",
                 "uncertain"),
      ordered = TRUE),
    trend_fine = factor(trend_fine,
      levels = c(
        "strong increase",
        "increase",
        "moderate increase",
        "stable",
        "moderate decrease",
        "decrease",
        "strong decrease",
        "uncertain"),
      ordered = TRUE)
  )
```

```{r}
summary_draw %>%
  bind_rows(extra_spec_df) %>%
  ggplot(aes(x = trend_coarse, fill = trend_coarse)) +
    geom_bar() +
    geom_text(aes(label = after_stat(count)),
              stat = "count",
              vjust = -0.5) +
    scale_x_discrete(drop = FALSE) +
    scale_fill_manual(values =  c("chartreuse3",
                                  "gold",
                                  "firebrick1",
                                  "skyblue"),
                      drop = FALSE) +
    theme(legend.position = "") +
    labs(y = "Number of species",
         x = "")
```

```{r}
summary_draw %>%
  bind_rows(extra_spec_df) %>%
  ggplot(aes(x = trend_fine, fill = trend_fine)) +
    geom_bar() +
    geom_text(aes(label = after_stat(count)),
              stat = "count",
              vjust = -0.5) +
    scale_x_discrete(drop = FALSE) +
    scale_fill_manual(values =  c(rev(traffic_palette(7)), "skyblue"),
                      drop = FALSE) +
    theme(legend.position = "",
          axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(y = "Number of species",
         x = "")
```


# Export results

We write out a .csv table with summary statistics for the research paper.
We round to four decimals.
We also include trends of the `r nrow(extra_spec_df)` who emerged or disappeared (zero counts).
They have `NA` values for the posterior median and credible intervals.

```{r}
organised_summary_df <- summary_draw %>%
  bind_rows(extra_spec_df) %>%
  select(species_nl, sci.median, sci.lower, sci.upper,
         trend_coarse, trend_fine, .point, .interval, .width) %>%
  mutate(across(starts_with("sci"), ~ round(.x, 4))) %>%
  arrange(species_nl)
```

```{r}
write_csv(organised_summary_df, "output/tables/sci_species_summary.csv")
```
