---
title: "Create figures for Dutch article"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Set up
library(knitr)
library(here)
opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(root.dir = here::here())

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)

# Packages
library(zen4R)
library(tidyverse)
library(INBOtheme)
theme_set(theme_inbo(transparent = TRUE))
```

# Goal

Create nice figures for a Dutch article in 'natuur.focus'.

```{r}
# Function to save figures using multiple devices
save_figure <- function(plot, file, devices, ...) {
  sapply(devices, function(dev) {
    ggplot2::ggsave(
      filename = paste0(file.path("media", file), ".", dev),
      plot = plot,
      ...
      )
  })
}
```

```{r}
# Devices to export
devices <- c("png")
```


# Load data

We load the data of the MSCI results.

```{r load-data}
# Create data directory if necessary
dir.create("output", showWarnings = FALSE, recursive = TRUE)

# Download data from zenodo if necessary
## MSCI
file_msci <- file.path("output", "msci_draw_list_log.rds")
if (file.exists(file_msci)) {
  msci_draw_list_log <- readRDS(file_msci)
} else {
  download_zenodo("10.5281/zenodo.8383838", path = "output",
                  files = list("msci_draw_list_log.rds"))
  msci_draw_list_log <- readRDS(file_msci)
}
```

```{r media-dir}
# create media directory if not yet available
dir.create("media", showWarnings = FALSE, recursive = TRUE)
```


# Set general functions and thresholds

We specify some general plot specifications as functions.

```{r}
index_labels <- function(x) {
  sprintf("%+.0f%%", 100 * (exp(x) - 1))
}

index_breaks <- function() {
  z <- 1 - c(
    4 / 5, 3 / 4, 2 / 3, 1 / 2, 1 / 3, 1 / 5
  )
  z <- log(sort(z))
  c(z, 0, -z)
}

minor_breaks <- function() {
  breaks <- unique(abs(index_breaks()))
  out <- vector(length = length(breaks) - 1)
  i <- 1
  while (i < length(breaks)) {
    out[i] <- mean(c(breaks[i], breaks[i + 1]))
    i <- i + 1
  }
  return(c(-out, out))
}

index_breaks_rev <- function() {
  exp(index_breaks())
}

index_labels_rev <- function(digits = 2) {
  format(round(exp(index_breaks()), digits), nsmall = digits)
}
```

We set thresholds at -20% and +25% in the log scale.

```{r}
min_threshold <- log(0.8)
max_threshold <- log(1.25)
```


# MSCI figures
## Biotope

We rename some trait values and sort them according to the posterior median.

```{r}
p_biotope <- msci_draw_list_log$Biotope %>%
  mutate(trait_value_new = case_when(
    trait_value == "WoodlandDeciduous" ~ "Deciduous woodland",
    trait_value == "WoodlandConiferous" ~ "Coniferous woodland",
    trait_value == "HeathlandWet" ~ "Wet heathland",
    trait_value == "HeathlandDry" ~ "Dry heathland",
    trait_value == "GrasslandNutrientPoor" ~ "Nutrient-poor grassland",
    trait_value == "GrasslandNutrientRich" ~ "Nutrient-rich grassland",
    trait_value == "Hedges & orchards" ~ "Hedges and orchards",
    TRUE ~ trait_value
    )
  ) %>%
  mutate(label = paste0(trait_value_new, "\n", "[n = ", n_species, "]"),
         label = reorder(label, msci.median,  decreasing = FALSE)) %>%
  ggplot(aes(y = label)) +
    geom_violin(aes(x = msci.draw, fill = trend_coarse),
                colour = alpha("white", 0)) +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
               linetype = "dotdash") +
    geom_errorbar(aes(xmin = msci.lower, xmax = msci.upper),
                  linewidth = 1, colour = "black", width = 0.1) +
    geom_point(aes(x = msci.median),
               size = 2.2, colour = "black") +
    scale_fill_manual(values =  c("chartreuse3",
                                  "gold",
                                  "firebrick1",
                                  "skyblue"),
                      drop = FALSE) +
    labs(y = "",
         x = "Multi-Species Change Index",
         fill = "Trend") +
    scale_x_continuous(
      breaks = index_breaks(),
      labels = index_labels,
      minor_breaks = minor_breaks(),
      limits = c(-log(3.3), log(3.3)),
      sec.axis = sec_axis(~ exp(.),
        breaks = index_breaks_rev(),
        labels = index_labels_rev(),
        name = "Proportional change")
      ) +
    theme(
      legend.position = c(1, 0),
      legend.justification = c(1, 0),
      legend.background = element_rect(fill = "white", color = "darkgrey"),
      legend.margin = margin(6, 6, 6, 6)
      )

p_biotope1 <- p_biotope +
  annotate("text", label = c("increase", "strong increase"),
           x = c(0 + 0.02, max_threshold + 0.02), y = c(0.3, 0.3),
           size = 2.5, hjust = "left") +
  annotate("segment",
           x = c(0 + 0.02, max_threshold + 0.02), y = c(0.4, 0.4),
           xend = c(0 + 0.1, max_threshold + 0.1), yend = c(0.4, 0.4),
           arrow = arrow(type = "closed", length = unit(0.005, "npc"))) +
  annotate("text", label = c("decrease", "strong decrease"),
           x = c(0 - 0.02, min_threshold - 0.02), y = c(0.3, 0.3),
           size = 2.5, hjust = "right") +
  annotate("segment",
           x = c(0 - 0.02, min_threshold - 0.02), y = c(0.4, 0.4),
           xend = c(0 - 0.1, min_threshold - 0.1), yend = c(0.4, 0.4),
           arrow = arrow(type = "closed", length = unit(0.005, "npc"))) +
  expand_limits(y = c(0.001, 0)) +
  theme(
    axis.text.x = element_text(size = 11),
    axis.text.y = element_text(size = 13),
    axis.title = element_text(size = 15)
    ) +
  theme(
    legend.key.size = unit(0.7, "cm"), # Change legend key size
    legend.key.height = unit(0.7, "cm"), # Change legend key height
    legend.key.width = unit(0.7, "cm"), # Change legend key width
    legend.title = element_text(size = 12), # Change legend title font size
    legend.text = element_text(size = 10) # Change legend text font size
    )

p_biotope1
```


```{r}
save_figure(p_biotope1, "msci_biotope_nl", devices,
            dpi = 300, width = 8, height = 8)
```


## Larval food source

We rename some trait values and sort them according to the posterior median.
We do not include trait value 'Diverse'.

```{r}
p_larvalfoodsource <- msci_draw_list_log$LarvalFoodSource %>%
  filter(trait_value != "Diverse") %>%
  mutate(trait_value_new = case_when(
    trait_value == "Shrubs_Trees" ~ "Shrubs and trees",
    trait_value == "Lichens_Mosses" ~ "Lichens and mosses",
    TRUE ~ trait_value
    )
  ) %>%
  mutate(label = paste0(trait_value_new, "\n", "[n = ", n_species, "]"),
         label = reorder(label, msci.median,  decreasing = FALSE)) %>%
  ggplot(aes(y = label)) +
    geom_violin(aes(x = msci.draw, fill = trend_coarse),
                colour = alpha("white", 0)) +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
               linetype = "dotdash") +
    geom_errorbar(aes(xmin = msci.lower, xmax = msci.upper),
                  linewidth = 1, colour = "black", width = 0.1) +
    geom_point(aes(x = msci.median),
               size = 2.2, colour = "black") +
    scale_fill_manual(values =  c("chartreuse3",
                                  "gold",
                                  "firebrick1",
                                  "skyblue"),
                      drop = FALSE) +
    labs(y = "",
         x = "Multi-Species Change Index",
         fill = "Trend") +
    scale_x_continuous(
      breaks = index_breaks(),
      labels = index_labels,
      minor_breaks = minor_breaks(),
      limits = c(-log(3.3), log(3.3)),
      sec.axis = sec_axis(~ exp(.),
        breaks = index_breaks_rev(),
        labels = index_labels_rev(),
        name = "Proportional change")
      ) +
    theme(
      legend.position = c(1, 0),
      legend.justification = c(1, 0),
      legend.background = element_rect(fill = "white", color = "darkgrey"),
      legend.margin = margin(6, 6, 6, 6)
      )

p_larvalfoodsource1 <- p_larvalfoodsource +
  annotate("text", label = c("increase", "strong increase"),
           x = c(0 + 0.02, max_threshold + 0.02), y = c(0.25, 0.25),
           size = 2.5, hjust = "left") +
  annotate("segment",
           x = c(0 + 0.02, max_threshold + 0.02), y = c(0.35, 0.35),
           xend = c(0 + 0.1, max_threshold + 0.1), yend = c(0.35, 0.35),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("text", label = c("decrease", "strong decrease"),
           x = c(0 - 0.02, min_threshold - 0.02), y = c(0.25, 0.25),
           size = 2.5, hjust = "right") +
  annotate("segment",
           x = c(0 - 0.02, min_threshold - 0.02), y = c(0.35, 0.35),
           xend = c(0 - 0.1, min_threshold - 0.1), yend = c(0.35, 0.35),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  expand_limits(y = c(0.001, 0)) +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14)
    ) +
  theme(
    legend.key.size = unit(0.5, "cm"), # Change legend key size
    legend.key.height = unit(0.5, "cm"), # Change legend key height
    legend.key.width = unit(0.5, "cm"), # Change legend key width
    legend.title = element_text(size = 10), # Change legend title font size
    legend.text = element_text(size = 8) # Change legend text font size
    )

p_larvalfoodsource1
```


```{r}
save_figure(p_larvalfoodsource1, "msci_larvalfoodsource_nl", devices,
            dpi = 300, width = 8, height = 4)
```


## Forewing length

We rename some trait values and sort them according to a logical order.

```{r}
trait_value_wing <- as.character(msci_draw_list_log$ForewingLength$trait_value)
trait_value_split <- strsplit(trait_value_wing, "(?<=[a-z])(?=[A-Z])",
                              perl = TRUE)
trait_value_wing_new <- sapply(trait_value_split, function(vec) {
  if (length(vec) == 2) {
    out <- paste(vec[1], tolower(vec[2]))
  } else {
    out <- vec
  }

  return(out)
  }
)
```

```{r}
p_forewinglength <- msci_draw_list_log$ForewingLength %>%
  mutate(trait_value_new = trait_value_wing_new) %>%
  arrange(trait_value) %>%
  rownames_to_column("order") %>%
  mutate(order = as.numeric(order)) %>%
  mutate(label = paste0(trait_value_new, "\n", "[n = ", n_species, "]"),
         label = reorder(label, order,  decreasing = TRUE)) %>%
  ggplot(aes(y = label)) +
    geom_violin(aes(x = msci.draw, fill = trend_coarse),
                colour = alpha("white", 0)) +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
               linetype = "dotdash") +
    geom_errorbar(aes(xmin = msci.lower, xmax = msci.upper),
                  linewidth = 1, colour = "black", width = 0.1) +
    geom_point(aes(x = msci.median),
               size = 2.2, colour = "black") +
    scale_fill_manual(values =  c("chartreuse3",
                                  "gold",
                                  "firebrick1",
                                  "skyblue"),
                      drop = FALSE) +
    labs(y = "",
         x = "Multi-Species Change Index",
         fill = "Trend") +
    scale_x_continuous(
      breaks = index_breaks(),
      labels = index_labels,
      minor_breaks = minor_breaks(),
      limits = c(-log(3.3), log(3.3)),
      sec.axis = sec_axis(~ exp(.),
        breaks = index_breaks_rev(),
        labels = index_labels_rev(),
        name = "Proportional change")
      ) +
    theme(
      legend.position = c(1, 0),
      legend.justification = c(1, 0),
      legend.background = element_rect(fill = "white", color = "darkgrey"),
      legend.margin = margin(6, 6, 6, 6)
      )

p_forewinglength1 <- p_forewinglength +
  annotate("text", label = c("increase", "strong increase"),
           x = c(0 + 0.02, max_threshold + 0.02), y = c(0.25, 0.25),
           size = 2.5, hjust = "left") +
  annotate("segment",
           x = c(0 + 0.02, max_threshold + 0.02), y = c(0.35, 0.35),
           xend = c(0 + 0.1, max_threshold + 0.1), yend = c(0.35, 0.35),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("text", label = c("decrease", "strong decrease"),
           x = c(0 - 0.02, min_threshold - 0.02), y = c(0.25, 0.25),
           size = 2.5, hjust = "right") +
  annotate("segment",
           x = c(0 - 0.02, min_threshold - 0.02), y = c(0.35, 0.35),
           xend = c(0 - 0.1, min_threshold - 0.1), yend = c(0.35, 0.35),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  expand_limits(y = c(0.001, 0)) +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14)
    ) +
  theme(
    legend.key.size = unit(0.5, "cm"), # Change legend key size
    legend.key.height = unit(0.5, "cm"), # Change legend key height
    legend.key.width = unit(0.5, "cm"), # Change legend key width
    legend.title = element_text(size = 10), # Change legend title font size
    legend.text = element_text(size = 8) # Change legend text font size
    )

p_forewinglength1
```


```{r}
save_figure(p_forewinglength1, "msci_forewinglength_nl", devices,
            dpi = 300, width = 8, height = 4)
```


## Host plant specificity

We sort the trait values them according to a logical order.

```{r}
p_hostplantspecificity <- msci_draw_list_log$HostPlantSpecificity %>%
  arrange(trait_value) %>%
  rownames_to_column("order") %>%
  mutate(order = as.numeric(order)) %>%
  mutate(label = paste0(trait_value, "\n", "[n = ", n_species, "]"),
         label = reorder(label, order,  decreasing = TRUE)) %>%
  ggplot(aes(y = label)) +
    geom_violin(aes(x = msci.draw, fill = trend_coarse),
                colour = alpha("white", 0)) +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
               linetype = "dotdash") +
    geom_errorbar(aes(xmin = msci.lower, xmax = msci.upper),
                  linewidth = 1, colour = "black", width = 0.1) +
    geom_point(aes(x = msci.median),
               size = 2.2, colour = "black") +
    scale_fill_manual(values =  c("chartreuse3",
                                  "gold",
                                  "firebrick1",
                                  "skyblue"),
                      drop = FALSE) +
    labs(y = "",
         x = "Multi-Species Change Index",
         fill = "Trend") +
    scale_x_continuous(
      breaks = index_breaks(),
      labels = index_labels,
      minor_breaks = minor_breaks(),
      limits = c(-log(3.3), log(3.3)),
      sec.axis = sec_axis(~ exp(.),
        breaks = index_breaks_rev(),
        labels = index_labels_rev(),
        name = "Proportional change")) +
    theme(
      legend.position = c(1, 0),
      legend.justification = c(1, 0),
      legend.background = element_rect(fill = "white", color = "darkgrey"),
      legend.margin = margin(6, 6, 6, 6)
      )

p_hostplantspecificity1 <- p_hostplantspecificity +
  annotate("text", label = c("increase", "strong increase"),
           x = c(0 + 0.02, max_threshold + 0.02), y = c(0.25, 0.25),
           size = 2.5, hjust = "left") +
  annotate("segment",
           x = c(0 + 0.02, max_threshold + 0.02), y = c(0.35, 0.35),
           xend = c(0 + 0.1, max_threshold + 0.1), yend = c(0.35, 0.35),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("text", label = c("decrease", "strong decrease"),
           x = c(0 - 0.02, min_threshold - 0.02), y = c(0.25, 0.25),
           size = 2.5, hjust = "right") +
  annotate("segment",
           x = c(0 - 0.02, min_threshold - 0.02), y = c(0.35, 0.35),
           xend = c(0 - 0.1, min_threshold - 0.1), yend = c(0.35, 0.35),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  expand_limits(y = c(0.001, 0)) +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14)
    ) +
  theme(
    legend.key.size = unit(0.5, "cm"), # Change legend key size
    legend.key.height = unit(0.5, "cm"), # Change legend key height
    legend.key.width = unit(0.5, "cm"), # Change legend key width
    legend.title = element_text(size = 10), # Change legend title font size
    legend.text = element_text(size = 8) # Change legend text font size
    )

p_hostplantspecificity1
```


```{r}
save_figure(p_hostplantspecificity1, "msci_hostplantspecificity_nl",
            dpi = 300, devices, width = 8, height = 4)
```


## Migration

We sort the trait values according to the posterior mean.

```{r}
p_migration <- msci_draw_list_log$Migration %>%
  mutate(label = paste0(trait_value, "\n", "[n = ", n_species, "]"),
         label = reorder(label, msci.median,  decreasing = FALSE)) %>%
  ggplot(aes(y = label)) +
    geom_violin(aes(x = msci.draw, fill = trend_coarse),
                colour = alpha("white", 0)) +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
               linetype = "dotdash") +
    geom_errorbar(aes(xmin = msci.lower, xmax = msci.upper),
                  linewidth = 1, colour = "black", width = 0.1) +
    geom_point(aes(x = msci.median),
               size = 2.2, colour = "black") +
    scale_fill_manual(values =  c("chartreuse3",
                                  "gold",
                                  "firebrick1",
                                  "skyblue"),
                      drop = FALSE) +
    labs(y = "",
         x = "Multi-Species Change Index",
         fill = "Trend") +
    scale_x_continuous(
      breaks = index_breaks(),
      labels = index_labels,
      minor_breaks = minor_breaks(),
      limits = c(-log(3.3), log(3.3)),
      sec.axis = sec_axis(~ exp(.),
        breaks = index_breaks_rev(),
        labels = index_labels_rev(),
        name = "Proportional change")
      ) +
    theme(
      legend.position = c(1, 0),
      legend.justification = c(1, 0),
      legend.background = element_rect(fill = "white", color = "darkgrey"),
      legend.margin = margin(6, 6, 6, 6)
      )

p_migration1 <- p_migration +
  annotate("text", label = c("increase", "strong increase"),
           x = c(0 + 0.02, max_threshold + 0.02), y = c(0.25, 0.25),
           size = 2.5, hjust = "left") +
  annotate("segment",
           x = c(0 + 0.02, max_threshold + 0.02), y = c(0.35, 0.35),
           xend = c(0 + 0.1, max_threshold + 0.1), yend = c(0.35, 0.35),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("text", label = c("decrease", "strong decrease"),
           x = c(0 - 0.02, min_threshold - 0.02), y = c(0.25, 0.25),
           size = 2.5, hjust = "right") +
  annotate("segment",
           x = c(0 - 0.02, min_threshold - 0.02), y = c(0.35, 0.35),
           xend = c(0 - 0.1, min_threshold - 0.1), yend = c(0.35, 0.35),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  expand_limits(y = c(0.001, 0)) +
  theme(
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 14)
    ) +
  theme(
    legend.key.size = unit(0.5, "cm"), # Change legend key size
    legend.key.height = unit(0.5, "cm"), # Change legend key height
    legend.key.width = unit(0.5, "cm"), # Change legend key width
    legend.title = element_text(size = 10), # Change legend title font size
    legend.text = element_text(size = 8) # Change legend text font size
    )

p_migration1
```


```{r}
save_figure(p_migration1, "msci_migration_nl", devices,
            dpi = 300, width = 8, height = 4)
```
