---
title: "Data exploration"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Set up
library(knitr)
library(here)
opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(root.dir = here::here())

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)
conflicted::conflicts_prefer(dplyr::first)

# Packages
library(zen4R)
library(tidyverse)
library(FactoMineR) # MCA
library(factoextra) # visualise MCA
library(corrplot)   # visualise correlations
library(gridExtra)
library(ggrepel)
```

# Goal

Exploration of the data. Where are the bottlenecks? Missing data? ...

# Load data

```{r load-data}
# Data path and create directory if necessary
data_path <- file.path("data", "intermediate")
dir.create(data_path, showWarnings = FALSE, recursive = TRUE)

# Download data from zenodo if necessary
file_prop_nozero <- file.path(data_path,
                              "prop_data_per_trait_nozero_intermed.csv")
if (file.exists(file_prop_nozero)) {
  species_traits_final_nozero <- read_csv(file_prop_nozero)
} else {
  download_zenodo("10.5281/zenodo.8383838", path = data_path,
                  files = list("prop_data_per_trait_nozero_intermed.csv"))
  species_traits_final_nozero <- read_csv(file_prop_nozero)
}

file_prop <- file.path(data_path, "prop_data_per_trait_intermed.csv")
if (file.exists(file_prop)) {
  species_traits_final <- read_csv(file_prop)
} else {
  download_zenodo("10.5281/zenodo.8383838", path = data_path,
                  files = list("prop_data_per_trait_intermed.csv"))
  species_traits_final <- read_csv(file_prop)
}
```

In total we have the following number of species.

```{r}
species_traits_final %>%
  select(species_nl, trait_name, trait_value) %>%
  filter(trait_name == "Migration") %>%
  distinct() %>%
  count(trait_value) %>%
  kable()
```

If we exclude species with zero counts:

```{r}
species_traits_final_nozero %>%
  select(species_nl, trait_name, trait_value) %>%
  filter(trait_name == "Migration") %>%
  distinct() %>%
  count(trait_value) %>%
  kable()
```


# Missing data and duplicates

We do not know all the traits for the following species.

```{r}
species_traits_final %>%
  filter(is.na(trait_value)) %>%
  distinct(species_nl, trait_name) %>%
  mutate(val = 1) %>%
  pivot_wider(names_from = trait_name, values_from = val) %>%
  mutate(across(-c(species_nl), ~ replace(.x, .x == 1, cur_column()))) %>%
  unite("trait_name", -c(species_nl), sep = ", ", na.rm = TRUE) %>%
  kable()
```

This is not an issue since we have the combined trait `TempHum`.

Some species have more than one trait value within the same trait.

```{r}
dubbles <- species_traits_final %>%
  count(species_nl, period, trait_name) %>%
  filter(n > 1) %>%
  distinct(species_nl, trait_name) %>%
  inner_join(
    distinct(
      species_traits_final, species_nl, trait_name,
      trait_value
    ),
    by = join_by(species_nl, trait_name)
  )

dubbles %>%
  kable()
```

We also calculate this for the data without zeroes.

```{r}
dubbles_nozero <- species_traits_final_nozero %>%
  count(species_nl, period, trait_name) %>%
  filter(n > 1) %>%
  distinct(species_nl, trait_name) %>%
  inner_join(
    distinct(
      species_traits_final, species_nl, trait_name,
      trait_value
    ),
    by = join_by(species_nl, trait_name)
  )
```

## Resolve issues

We avoid problems for species with more than one trait value within the same trait by giving a suffix number to the species in a separate column. By grouping the data, it is possible to circumvent any issues (see code further on).

We remove traits `HabitatTemperature` and `HabitatHumidity` for further processing.
We have no `NA`'s left.

```{r}
design_df <- dubbles %>%
  rowid_to_column("id") %>%
  mutate(species_new = ifelse(id %% 2 != 0,
    paste(species_nl, 1, sep = "_"),
    paste(species_nl, 2, sep = "_")
  )) %>%
  select(-id)

species_traits_final_new <- species_traits_final %>%
  full_join(design_df, by = join_by(species_nl, trait_name, trait_value)) %>%
  mutate(
    species_new = ifelse(!is.na(species_new), species_new,
      ifelse(species_nl %in% unique(design_df$species_nl),
        paste(species_nl, 1, sep = "_"), species_nl
      )
    )
  ) %>%
  filter(!trait_name %in% c("HabitatTemperature", "HabitatHumidity"))
```

We also calculate this for the data without zeroes.

```{r}
design_df_nozero <- dubbles_nozero %>%
  rowid_to_column("id") %>%
  mutate(species_new = ifelse(id %% 2 != 0,
    paste(species_nl, 1, sep = "_"),
    paste(species_nl, 2, sep = "_")
  )) %>%
  select(-id)

spec_traits_final_nozero_new <- species_traits_final_nozero %>%
  full_join(design_df_nozero,
            by = join_by(species_nl, trait_name, trait_value)) %>%
  mutate(
    species_new = ifelse(!is.na(species_new), species_new,
      ifelse(species_nl %in% unique(design_df$species_nl),
        paste(species_nl, 1, sep = "_"), species_nl
      )
    )
  ) %>%
  filter(!trait_name %in% c("HabitatTemperature", "HabitatHumidity"))
```

# Summaries
## Univariate

How many data do we have for each trait (for data including zeroes)?

```{r}
univar_summary <- species_traits_final_new %>%
  count(trait_name, trait_value, period) %>%
  distinct(trait_name, trait_value, n) %>%
  mutate("n <= 10" = ifelse(n <= 10, "YES", "no")) %>%
  arrange(trait_name, trait_value)

univar_summary %>%
  kable()
```

We see some trait values with less than 10 species.

```{r}
table(univar_summary$`n <= 10`)
```

## Bivariate

We can look at combinations of traits. Do we see any interesting correlations?

```{r}
cross_tab <- function(df, row, column, print = TRUE) {
  require("dplyr")
  require("tidyr")
  require("data.table")
  cross_table <- df %>%
    filter(.data$trait_name %in% c(row, column)) %>%
    distinct(
      .data$species_nl, .data$species_new, .data$trait_name,
      .data$trait_value
    ) %>%
    pivot_wider(
      id_cols = c("species_nl", "species_new"), names_from = .data$trait_name,
      values_from = .data$trait_value
    ) %>%
    # Fill in missing trait values for species that occur twice
    group_by(.data$species_nl) %>%
    arrange(.data$species_nl, .data$species_new, !!sym(row), !!sym(column)) %>%
    mutate(
      "{row}" := ifelse(is.na(!!sym(row)), # nolint: object_name_linter.
        first(!!sym(row)), !!sym(row)
      ),
      "{column}" := ifelse(is.na(!!sym(column)), # nolint: object_name_linter.
        first(!!sym(column)), !!sym(column)
      )
    ) %>%
    ungroup() %>%
    # Pivot wider can recreate NA's for missing trait values
    # (which were filtered out before in the dataset)
    drop_na() %>%
    # How many species per trait combination?
    count(!!sym(row), !!sym(column)) %>%
    pivot_wider(
      id_cols = !!sym(row), names_from = !!sym(column),
      values_from = .data$n
    ) %>%
    mutate_all(~ replace(., is.na(.), 0))

  if (isTRUE(print)) {
    return(print(cross_table))
  } else {
    return(cross_table)
  }
}
```

```{r}
small_traits <- univar_summary %>%
  filter(n <= 10) %>%
  pull(trait_value)

filtered_df <- species_traits_final_new %>%
  filter(!trait_value %in% small_traits,
         trait_name != "Migration")

combinations <- combn(unique(filtered_df$trait_name), 2)

for (i in seq_len(ncol(combinations))) {
  combo <- combinations[, i]

  print("------------------------------------------------")
  print(combo)
  cross_tab(filtered_df, combo[1], combo[2])
}
```

## Multivariate

Multiple Correspondence Analysis ([MCA](http://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/114-mca-multiple-correspondence-analysis-in-r-essentials/)) is an alternative to PCA for categorical variables. The `MCA()` function from the **FactoMineR** package can be used to implement MCA.
We do not include `Family`.

```{r}
dat_mca <- filtered_df %>%
  filter(trait_name != "Family") %>%
  select(species_new, trait_name, trait_value) %>%
  distinct() %>%
  pivot_wider(
    id_cols = species_new,
    names_from = trait_name, values_from = trait_value
  ) %>%
  select(-species_new) %>%
  drop_na()

traits_mca <- MCA(dat_mca, graph = FALSE)
```

We visualise the percentages of variance explained by each of the MCA dimensions.

```{r}
fviz_screeplot(traits_mca, addlabels = TRUE, ylim = c(0, 10))
```

We visualise the correlation between variables and MCA principal dimensions.

```{r}
# bug in fviz_mca_var function of factoextrapackage
# https://github.com/kassambara/factoextra/issues/166
# create own plots ...

plot_df <- as.data.frame(traits_mca$var$eta2) %>%
  rownames_to_column("variable")

p1 <- ggplot(plot_df, aes(x = `Dim 1`, y = `Dim 2`)) +
    geom_hline(yintercept = 0, colour = "grey", linetype = "dashed") +
    geom_vline(xintercept = 0, colour = "grey", linetype = "dashed") +
    geom_point() +
    geom_text_repel(aes(label = variable), size = 3) +
    coord_cartesian(xlim = c(-0.05, NA), ylim = c(-0.05, NA)) +
    theme_minimal()

p2 <- ggplot(plot_df, aes(x = `Dim 2`, y = `Dim 3`)) +
    geom_hline(yintercept = 0, colour = "grey", linetype = "dashed") +
    geom_vline(xintercept = 0, colour = "grey", linetype = "dashed") +
    geom_point() +
    geom_text_repel(aes(label = variable), size = 3) +
    coord_cartesian(xlim = c(-0.05, NA), ylim = c(-0.05, NA)) +
    theme_minimal()

p3 <- ggplot(plot_df, aes(x = `Dim 3`, y = `Dim 4`)) +
    geom_hline(yintercept = 0, colour = "grey", linetype = "dashed") +
    geom_vline(xintercept = 0, colour = "grey", linetype = "dashed") +
    geom_point() +
    geom_text_repel(aes(label = variable), size = 3) +
    coord_cartesian(xlim = c(-0.05, NA), ylim = c(-0.05, NA)) +
    theme_minimal()

p4 <- ggplot(plot_df, aes(x = `Dim 4`, y = `Dim 5`)) +
    geom_hline(yintercept = 0, colour = "grey", linetype = "dashed") +
    geom_vline(xintercept = 0, colour = "grey", linetype = "dashed") +
    geom_point() +
    geom_text_repel(aes(label = variable), size = 3) +
    coord_cartesian(xlim = c(-0.05, NA), ylim = c(-0.05, NA)) +
    theme_minimal()

grid.arrange(p1, p2, p3, p4)
```

Variable categories with a similar species profile are grouped together. The distance between category points and the origin measures the quality of the variable category on the factor map. Category points that are away from the origin are well represented on the factor map. The two dimensions 1 and 2 are sufficient to retain 11.4 % of the total variation contained in the data. Not all the points are equally well displayed in these two dimensions.The quality of the representation is called the squared cosine (cos2), which measures the degree of association between variable categories and a particular axis

```{r}
fviz_mca_var(traits_mca, axes = c(1, 2), repel = TRUE,
             col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             ggtheme = theme_minimal())

fviz_mca_var(traits_mca, axes = c(3, 4), repel = TRUE,
             col.var = "cos2",
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             ggtheme = theme_minimal())
```

```{r}
p1 <- fviz_contrib(traits_mca, choice = "var", axes = 1, top = 15,
                   font.xtickslab = 7, font.title = 10)
p2 <- fviz_contrib(traits_mca, choice = "var", axes = 2, top = 15,
                   font.xtickslab = 7, font.title = 10)
p3 <- fviz_contrib(traits_mca, choice = "var", axes = 3, top = 15,
                   font.xtickslab = 7, font.title = 10)
p4 <- fviz_contrib(traits_mca, choice = "var", axes = 4, top = 15,
                   font.xtickslab = 7, font.title = 10)
grid.arrange(p1, p2, p3, p4)
```

Along all dimensions:

```{r, fig.height=10}
var <- get_mca_var(traits_mca)
corrplot(var$cos2, is.corr = FALSE, tl.col = "black", tl.srt = 45, tl.cex = 0.5,
         cl.pos = "b", cl.cex = 0.5, cl.length = 4)
```

```{r}
fviz_cos2(traits_mca, choice = "var", axes = 1:5)
```


# Geometric means
## Separate traits

We calculate the geometric mean and 95 % confidence interval based on 1000 bootstrap samples for all traits with more than 10 species. This is based on the data without zeroes.

```{r}
univar_summary_nozero <- spec_traits_final_nozero_new %>%
  count(trait_name, trait_value, period) %>%
  distinct(trait_name, trait_value, n) %>%
  mutate("n <= 10" = ifelse(n <= 10, "YES", "no")) %>%
  arrange(trait_name, trait_value)

small_traits_nozero <- univar_summary_nozero %>%
  filter(n <= 10) %>%
  pull(trait_value)

filtered_df_nozero <- spec_traits_final_nozero_new %>%
  filter(!trait_value %in% small_traits,
         trait_name != "Migration")

combinations_nozero <- combn(unique(filtered_df$trait_name), 2)
```

```{r}
geometric_mean <- filtered_df_nozero %>%
  select(-c(sum_per_period, n_grids)) %>%
  pivot_wider(names_from = period, values_from = rel_prop) %>%
  filter(p1980_2012 != 0 & p2013_2022 != 0) %>%
  mutate(
    sci = p2013_2022 / p1980_2012,
    log_sci = log(sci)
  ) %>%
  group_by(trait_name, trait_value) %>%
  summarise(
    mean_cl_boot(log_sci,
      conf.int = .95
    ),
    n_species = n()
  ) %>%
  mutate(
    msci = exp(y) - 1,
    msci_lwr = exp(ymin) - 1,
    msci_upr = exp(ymax) - 1
  )

geometric_mean %>%
  select(-c(y, ymin, ymax)) %>%
  arrange(trait_name, trait_value) %>%
  kable()
```

```{r}
traits <- sort(unique(geometric_mean$trait_name))

for (t in traits) {
  p <- geometric_mean %>%
    filter(trait_name == t) %>%
    mutate(col_var = case_when(
      msci_lwr > 0 ~ "chartreuse3",
      msci_upr < 0 ~ "coral",
      msci_lwr < 0 & msci_lwr > -0.2 & msci_upr > 0 & msci_upr < 0.25 ~ "gold",
      TRUE ~ "skyblue",
    )) %>%
    ggplot(aes(y = msci, x = trait_value)) +
    geom_hline(yintercept = 0, linetype = 2) +
    geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
    geom_point(aes(colour = col_var),
      size = 2,
      position = position_dodge(width = 0.6)
    ) +
    geom_errorbar(aes(ymin = msci_lwr, ymax = msci_upr, colour = col_var),
      width = 0.25, position = position_dodge(width = 0.6)
    ) +
    geom_text(aes(y = msci_upr, label = n_species, group = trait_value),
      vjust = -1,
      colour = "firebrick", position = position_dodge(width = 0.6)
    ) +
    scale_y_continuous(expand = expansion(mult = 0.2)) +
    scale_colour_identity() +
    labs(x = "", y = "Multispecies Change Index", title = t) +
    theme(legend.position = "")
  print(p)
}
```

## Combinations of traits

```{r}
for (i in seq_len(ncol(combinations_nozero))) {
  combo <- combinations_nozero[, i]

  if (!(setequal(c("HabitatTemperature", "TempHum"), combo) ||
    setequal(c("TempHum", "HabitatHumidity"), combo))) {
    row <- combo[1]
    column <- combo[2]
    combo_frame <- filtered_df_nozero %>%
      filter(trait_name %in% combo) %>%
      pivot_wider(
        id_cols = c(
          species_nl, species_new, n_grids, period, sum_per_period,
          rel_prop
        ),
        names_from = trait_name, values_from = trait_value
      ) %>%
      # Fill in missing trait values for species that occur twice
      group_by(species_nl) %>%
      arrange(species_nl, species_new, !!sym(row), !!sym(column)) %>%
      mutate(
        "{row}" := ifelse(is.na(!!sym(row)), # nolint: object_name_linter.
          first(!!sym(row)), !!sym(row)
        ),
        "{column}" := ifelse(is.na(!!sym(column)), # nolint: object_name_linter.
          first(!!sym(column)), !!sym(column)
        )
      ) %>%
      ungroup() %>%
      # Pivot wider can recreate NA's for missing trait values
      # (which were filtered out before in the dataset)
      drop_na() %>%
      # How many species per trait combination?
      group_by_at(c("period", combo)) %>%
      mutate(n = n()) %>%
      ungroup() %>%
      filter(
        n > 10
      ) %>%
      select(-c(sum_per_period, n_grids)) %>%
      pivot_wider(names_from = period, values_from = rel_prop) %>%
      filter(p1980_2012 != 0 & p2013_2022 != 0)

    combo_geometric_mean <- combo_frame %>%
      mutate(
        sci = p2013_2022 / p1980_2012,
        log_sci = log(sci)
      ) %>%
      group_by_at(c(combo, "n")) %>%
      summarise(
        mean_cl_boot(log_sci,
          conf.int = .95
        ),
        .groups = "drop"
      ) %>%
      mutate(
        msci = exp(y) - 1,
        msci_lwr = exp(ymin) - 1,
        msci_upr = exp(ymax) - 1
      )

    p <- combo_geometric_mean %>%
      mutate(
        col_var = case_when(
          msci_lwr > 0 ~ "chartreuse3",
          msci_upr < 0 ~ "coral",
          msci_lwr < 0 & msci_lwr > -0.2 & msci_upr > 0 & msci_upr < 0.25
          ~ "gold",
          TRUE ~ "skyblue",
        ),
        trait_value = paste(.[[combo[1]]], .[[combo[2]]], sep = "\n")
      ) %>%
      ggplot(aes(y = msci, x = .data[[combo[1]]])) +
      geom_hline(yintercept = 0, linetype = 2) +
      geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
      geom_point(aes(colour = col_var),
        size = 2,
        position = position_dodge(width = 0.6)
      ) +
      geom_errorbar(aes(ymin = msci_lwr, ymax = msci_upr, colour = col_var),
        width = 0.25, position = position_dodge(width = 0.6)
      ) +
      geom_text(aes(y = msci_upr, label = n, group = trait_value),
        vjust = -1, colour = "firebrick",
        position = position_dodge(width = 0.6)
      ) +
      scale_y_continuous(expand = expansion(mult = 0.2)) +
      scale_colour_identity() +
      labs(
        x = "", y = "Multispecies Change Index",
        title = paste(combo, collapse = " - ")
      ) +
      theme(
        legend.position = "",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
      ) +
      facet_wrap(~ .data[[combo[2]]])
    print(p)
  }
}
```


# Write out data

We write out the final datasets (for zero and non-zero datasets) with extra columns `species_new` for species occurring twice in a single trait group and a column `n_spec` containing the number of species per trait value.

```{r}
out_df <- species_traits_final_new %>%
  select(species_nl, species_new, everything()) %>%
  group_by(period, trait_value) %>%
  mutate(n_spec = n_distinct(species_new)) %>%
  ungroup() %>%
  arrange(species_nl, species_new, trait_name, trait_value)

out_df_nozero <- spec_traits_final_nozero_new %>%
  select(species_nl, species_new, everything()) %>%
  group_by(period, trait_value) %>%
  mutate(n_spec = n_distinct(species_new)) %>%
  ungroup() %>%
  arrange(species_nl, species_new, trait_name, trait_value)
```

```{r data-dir}
out_path <- file.path("data", "processed")

# create data directory if not yet available
dir.create(out_path, showWarnings = FALSE, recursive = TRUE)
```

```{r}
write_csv(
  out_df_nozero,
  file.path(out_path, "prop_data_per_trait_nozero.csv")
)
write_csv(
  out_df,
  file.path(out_path, "prop_data_per_trait.csv")
)
```
