---
title: "Data preparation"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
bibliography: [references.json]
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/research-institute-for-nature-and-forest.csl
---

```{r setup, include=FALSE}
# Set up
library(knitr)
library(here)
opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(root.dir = here::here())

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)

# Packages
library(tidyverse)
```

# Goal

Process and merge data to use for modelling. Write out final dataframe(s).

# Load data

We read in the data and prepare it for further processing. We select stationary and migratory species.

```{r}
data_path <- file.path("data", "raw")

# Species list
check_list_flanders <- read_delim(
  file.path(data_path, "tblChecklistFlanders.csv"),
  delim = ";"
) %>%
  filter(Status == "Standvlinder" | Status == "Trekvlinder") %>%
  select(species_nl)

# Presences
presence_data <- read_delim(
  file.path(data_path, "tblAllCondensUtm5km_v2022.csv"),
  delim = ";"
) %>%
  rename(
    id = Id,
    utm_5km = Utm5km
  )
```

# Data processing

We merge the species list and presence list and count the number of occupied UTM grids per species per period. We store both a dataset with and without zero counts.

```{r}
# Join dataframes
joined_df <- left_join(check_list_flanders, presence_data, by = "species_nl")

# Number of grids per species per period
df_long <- joined_df %>%
  distinct(
    species_nl,
    year,
    utm_5km
  ) %>%
  filter(year >= 1980 & year <= 2022) %>%
  mutate(period = case_when(
    year >= 1980 & year <= 2012 ~ "p1980_2012",
    year >= 2013 & year <= 2022 ~ "p2013_2022"
  )) %>%
  distinct(
    species_nl,
    utm_5km,
    period
  ) %>%
  group_by(period) %>%
  mutate(sum_per_period = n()) %>%
  group_by(species_nl, period, sum_per_period) %>%
  summarise(n_grids = n())

# Save number of grids
number_of_grids <- df_long %>%
  ungroup() %>%
  distinct(period, sum_per_period)

# Wide format
df_wide <- df_long %>%
  pivot_wider(
    id_cols = species_nl,
    names_from = period,
    values_from = n_grids
  )

df_wide[c("p1980_2012", "p2013_2022")][is.na(df_wide[c(
  "p1980_2012",
  "p2013_2022"
)])] <- 0

# Save dataframe without zeroes
df_wide_sum_nozero <- df_wide %>%
  filter(p1980_2012 > 0 & p2013_2022 > 0) %>%
  mutate(sum = p1980_2012 + p2013_2022)

# Save dataframe with zeroes
df_wide_sum <- df_wide %>%
  mutate(sum = p1980_2012 + p2013_2022)
```

```{r}
# Visualise
head(df_wide_sum, 10) %>%
  kable()
```

We calculate the relative proportion $p$ of species $i$ in period $j$ as the number of UTM grids $n$ per species ($i = 1, ..., S$) in each period ($j = 1, 2$) divided by the sum of the total number of grids per period, namely `r pull(number_of_grids[1, 2])` in the reference period 1980-2012 ($j = 1$) and `r pull(number_of_grids[2, 2])` in the period 2013-2022 ($j = 2$).

<!-- spell-check: ignore:start -->
$$
\hat{p}_{ij} = \frac{n_{ij}}{\sum_{i=1}^S n_{ij}}
$$
<!-- spell-check: ignore:end -->

```{r}
# Without zeroes
prop_df_nozero <- df_wide_sum_nozero %>%
  mutate(p2013_2022 = ifelse(p1980_2012 == 0 & p2013_2022 == 0, NA,
    p2013_2022
  )) %>%
  pivot_longer(
    cols = c(p1980_2012:p2013_2022),
    names_to = "period",
    values_to = "n_grids"
  ) %>%
  group_by(period) %>%
  filter(!is.na(n_grids)) %>%
  mutate(
    sum_per_period = sum(n_grids),
    rel_prop = (n_grids) / sum_per_period
  ) %>%
  ungroup() %>%
  select(species_nl, n_grids, period, sum_per_period, rel_prop)

# With zeroes
prop_df <- df_wide_sum %>%
  mutate(p2013_2022 = ifelse(p1980_2012 == 0 & p2013_2022 == 0, NA,
    p2013_2022
  )) %>%
  pivot_longer(
    cols = c(p1980_2012:p2013_2022),
    names_to = "period",
    values_to = "n_grids"
  ) %>%
  group_by(period) %>%
  filter(!is.na(n_grids)) %>%
  mutate(
    sum_per_period = sum(n_grids),
    rel_prop = (n_grids) / sum_per_period
  ) %>%
  ungroup() %>%
  select(species_nl, n_grids, period, sum_per_period, rel_prop)
```

```{r}
# Visualise
head(prop_df, 10) %>%
  kable()
```

# Add species traits

We add species traits to the dataframes.

```{r}
# Read in data
species_traits <- read_delim(
  file.path(data_path, "tblMacromoths_Traits_v2022.csv"),
  delim = ";"
) %>%
  rename(species_nl = nameDutch)

species_traits$EllenbergN <- as.factor(species_traits$EllenbergN)
species_traits$nGenerations <- as.factor(species_traits$nGenerations)

# Biotope is in it 2 times, so we have to split up
species_traits_df1 <- species_traits %>%
  select(
    species_nl,
    Biotope1,
    colourVariation,
    Activity,
    OverwinteringStage,
    Phagy,
    LarvalFoodSource2,
    EllenbergN,
    nGenerations,
    Seasonality,
    WingDevelopment,
    Distribution,
    Migration,
    Size,
    TempHum,
    HabitatTemperature,
    HabitatHumidity
  ) %>%
  rename(Biotope = Biotope1) %>%
  filter(!is.na(species_nl))

species_traits_df2 <- species_traits %>%
  filter(!is.na(Biotope2))

species_traits_df2 <- species_traits_df2 %>%
  select(
    species_nl,
    Biotope2,
    colourVariation,
    Activity,
    OverwinteringStage,
    Phagy,
    LarvalFoodSource2,
    EllenbergN,
    nGenerations,
    Seasonality,
    WingDevelopment,
    Distribution,
    Migration,
    Size,
    TempHum,
    HabitatTemperature,
    HabitatHumidity
  ) %>%
  rename(Biotope = Biotope2)

species_traits_complete <- rbind(
  species_traits_df1,
  species_traits_df2
)

# Without zeroes
species_traits_joined_nozero <- prop_df_nozero %>%
  inner_join(species_traits_complete,
    by = "species_nl",
    relationship = "many-to-many"
  )

species_traits_final_nozero <- species_traits_joined_nozero %>%
  pivot_longer(
    cols = Biotope:HabitatHumidity,
    names_to = "trait_name",
    values_to = "trait_value",
    values_transform = as.character
  ) %>%
  distinct()

# With zeroes
species_traits_joined <- prop_df %>%
  inner_join(species_traits_complete,
    by = "species_nl",
    relationship = "many-to-many"
  )

species_traits_final <- species_traits_joined %>%
  pivot_longer(
    cols = Biotope:HabitatHumidity,
    names_to = "trait_name",
    values_to = "trait_value",
    values_transform = as.character
  ) %>%
  distinct()
```

```{r}
# Visualise
head(species_traits_final, 10) %>%
  kable()
```

We do not know all the traits for the following species.

```{r}
species_traits_final %>%
  filter(is.na(trait_value)) %>%
  distinct(species_nl, trait_name) %>%
  mutate(val = 1) %>%
  pivot_wider(names_from = trait_name, values_from = val) %>%
  mutate(across(-c(species_nl), ~ replace(.x, .x == 1, cur_column()))) %>%
  unite("trait_name", -c(species_nl), sep = ", ", na.rm = TRUE) %>%
  kable()
```

# Write out data

We write out the datasets.

```{r}
out_path <- file.path("data", "processed")

write_csv(
  species_traits_final_nozero,
  file.path(out_path, "prop_data_per_trait_nozero.csv")
)
write_csv(
  species_traits_final,
  file.path(out_path, "prop_data_per_trait.csv")
)
```
