---
title: "Data exploration"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
bibliography: [references.json]
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/research-institute-for-nature-and-forest.csl
---

```{r setup, include=FALSE}
# Set up
library(knitr)
library(here)
opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(root.dir = here::here())

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)

# Packages
library(tidyverse)
```

# Goal

Exploration of the data. Where are the bottlenecks? Missing data? ...

# Load data

```{r}
out_path <- file.path("data", "processed")

species_traits_final_nozero <- read_csv(
  file.path(out_path, "prop_data_per_trait_nozero.csv")
)
species_traits_final <- read_csv(file.path(out_path, "prop_data_per_trait.csv"))
```

# Missing data and duplicates

We do not know all the traits for the following species.

```{r}
species_traits_final %>%
  filter(is.na(trait_value)) %>%
  distinct(species_nl, trait_name) %>%
  mutate(val = 1) %>%
  pivot_wider(names_from = trait_name, values_from = val) %>%
  mutate(across(-c(species_nl), ~ replace(.x, .x == 1, cur_column()))) %>%
  unite("trait_name", -c(species_nl), sep = ", ", na.rm = TRUE) %>%
  kable()
```

Some species have more than one trait value within the same trait.

```{r}
dubbles <- species_traits_final %>%
  count(species_nl, period, trait_name) %>%
  filter(n > 1) %>%
  distinct(species_nl, trait_name) %>%
  inner_join(
    distinct(
      species_traits_final, species_nl, trait_name,
      trait_value
    ),
    by = join_by(species_nl, trait_name)
  )

dubbles %>%
  kable()
```

## Resolve issues

We avoid problems for species with more than one trait value within the same trait by giving a suffix number to the species. We also filter out `NA`'s.

```{r}
design_df <- dubbles %>%
  rowid_to_column("id") %>%
  mutate(species_new = ifelse(id %% 2 != 0,
    paste(species_nl, 1),
    paste(species_nl, 2)
  )) %>%
  select(-id)

species_traits_final_new <- species_traits_final %>%
  full_join(design_df, by = join_by(species_nl, trait_name, trait_value)) %>%
  mutate(species_nl = ifelse(!is.na(species_new),
    species_new,
    species_nl
  )) %>%
  select(-species_new) %>%
  filter(!is.na(trait_value))
```


# Summaries
## Univariate

How many data do we have for each trait?

```{r}
univar_summary <- species_traits_final_new %>%
  count(trait_name, trait_value, period) %>%
  distinct(trait_name, trait_value, n) %>%
  mutate("n <= 10" = ifelse(is.na(trait_value), "NA",
    ifelse(n <= 10, "YES", "no")
  )) %>%
  arrange(trait_name, trait_value)

univar_summary %>%
  kable()
```

We see some trait values with less than 10 species.

```{r}
table(univar_summary$`n <= 10`)
```

## Bivariate

We can look at combinations of traits. Do we see any interesting correlations?

```{r}
cross_tab <- function(df, row, column, print = TRUE) {
  cross_table <- df %>%
    filter(.data$trait_name %in% c(row, column)) %>%
    distinct(.data$species_nl, .data$trait_name, .data$trait_value) %>%
    pivot_wider(
      id_cols = .data$species_nl, names_from = .data$trait_name,
      values_from = .data$trait_value
    ) %>%
    drop_na() %>%
    count(!!sym(row), !!sym(column)) %>%
    pivot_wider(
      id_cols = !!sym(row), names_from = !!sym(column),
      values_from = .data$n
    ) %>%
    mutate_all(~ replace(., is.na(.), 0))

  if (isTRUE(print)) {
    return(print(cross_table))
  } else {
    return(cross_table)
  }
}
```

```{r}
small_traits <- univar_summary %>%
  filter(n <= 10) %>%
  pull(trait_value)

filtered_df <- species_traits_final_new %>%
  filter(!trait_value %in% small_traits)

combinations <- combn(unique(filtered_df$trait_name), 2)

for (i in seq_len(ncol(combinations))) {
  combo <- combinations[, i]

  if (!(setequal(c("HabitatTemperature", "TempHum"), combo) ||
    setequal(c("TempHum", "HabitatHumidity"), combo))) {
    print("------------------------------------------------")
    print(combo)
    cross_tab(filtered_df, combo[1], combo[2])
  }
}
```

# Geometric means

We calculate the geometric mean and 95 % confidence interval based on 1000 bootstrap samples for all traits with more than 10 species.

```{r}
geometric_mean <- species_traits_final_new %>%
  select(-c(sum_per_period, n_grids)) %>%
  pivot_wider(names_from = period, values_from = rel_prop) %>%
  filter(p1980_2012 != 0 & p2013_2022 != 0) %>%
  mutate(sci = p2013_2022 / p1980_2012,
         log_sci = log(sci)) %>%
  group_by(trait_name, trait_value) %>%
  summarise(mean_cl_boot(log_sci,
                         conf.int = .95)) %>%
  mutate(msci = exp(y) - 1,
         msci_lwr = exp(ymin) - 1,
         msci_upr = exp(ymax) - 1)

geometric_mean %>%
  select(-c(y, ymin, ymax)) %>%
  arrange(trait_name, trait_value) %>%
  kable()
```


