---
title: "Model specification and tests"
author: "Ward Langeraert"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    code_folding: hide
    toc: true
    toc_float: true
    toc_collapsed: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
# Set up
library(knitr)
library(here)
opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
opts_knit$set(root.dir = here::here())

conflicted::conflicts_prefer(dplyr::filter)
conflicted::conflicts_prefer(dplyr::select)
conflicted::conflicts_prefer(stats::lag)
conflicted::conflicts_prefer(brms::ar)
conflicted::conflicts_prefer(brms::dstudent_t)
conflicted::conflicts_prefer(brms::pstudent_t)
conflicted::conflicts_prefer(brms::qstudent_t)
conflicted::conflicts_prefer(brms::rstudent_t)

# Packages
library(tidyverse)
library(brms)
library(tidybayes)

# Source
source(here("source", "summarise_brms_models.R"))
```

# Goal

Test out model specifications and model fit.

# Load data

```{r}
data_path <- file.path("data", "processed")

species_traits_final_nozero <- read_csv(
  file.path(data_path, "prop_data_per_trait_nozero.csv")
)
species_traits_final <- read_csv(file.path(data_path,
                                           "prop_data_per_trait.csv"))
```

# Model specification

We assume a multinomial distribution per period $j$ ($j = 1, 2$) for the number of grid squares $n$ per species $i$ ($i = 1, ..., S$).

$$
\{n_{1j}, ..., n_{Sj}\} \sim Multinom(\sum_{i=1}^S n_{ij}, \{p_{1j}, ..., p_{Sj}\})
$$

Where $p_{ij} = \frac{\lambda_{ij}}{\lambda_{1j} + ... + \lambda_{Sj}}$

$\{n_{1j}, ..., n_{Sj}\}$ can be interpreted as independent Poisson random variables

$$
n_{ij} \sim Poisson(\lambda_{ij})
$$

Such that we can specify the model

$$
\ln(\lambda_{ij}) = \beta_0 + \beta_{msci}X_{period} + \ln(\Sigma_{i=1}^Sn_{ij}) + b_{0,{i = 1, ..., S}} + b_{1,{i = 1, ..., S}}X_{period}
$$

Where

-   $\beta_0$ the average log-number of grids for period 1
-   $\beta_{msci}$ the effect of period 2 compared to period 1 such that $MSCI = \exp(\beta_{msci})$
-   $X_{period}$ is 0 for period 1980-2012 and 1 for period 2013-2022
-   $ln(\Sigma_{i=1}^Sn_{ij})$ an offset used to control for the disproportionate search effort between the two periods
-   $b_{0,{i = 1, ..., S}}$ a random intercept for each species: deviation per species from $\beta_0$
-   $b_{1,{i = 1, ..., S}}$ a random slope for each species: deviation per species from $\beta_{msci}$

We set $corr(b_0, b_1) = 0$ since we want a separate (uncorrelated) deviation for both periods.

## Testing model structure

Example for trait 'Photoperiod'. First for trait value 'DayActive'.

```{r}
# create cache directory if not yet available
dir.create(here("source", "brms_cache"), FALSE)
```

```{r}
dat <- species_traits_final %>%
  filter(trait_value == "DayActive")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 4000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing

fit <- brm(
  bf(n_grids ~ period + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  file = "source/brms_cache/fit",
  file_refit = "on_change"
)
```

Convergence ok. 

```{r}
plot(fit)
```

Model fit ok.

```{r}
pp_check(fit,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "period"
)
```

Results:

```{r}
msci_draws_df1 <- fit %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  mutate(DayActive = exp(b_periodp2013_2022) - 1) %>%
  select(DayActive) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci")

summ_df1 <- msci_draws_df1 %>%
  group_by(trait) %>%
  mean_qi() %>%
  rename(mean = msci)

summ_df1 %>%
  kable()
```

```{r}
msci_draws_df1 %>%
  full_join(summ_df1, by = join_by(trait)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  ggplot(aes(y = msci, x = trait, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(legend.position = "")
```

We also create a single model for all trait values in 'Photoperiod'.

```{r}
dat2 <- species_traits_final %>%
  filter(trait_name == "Photoperiod")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit2 <- brm(
  bf(n_grids ~ period * trait_value + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat2, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit2",
  file_refit = "on_change"
)
```

Convergence ok.

```{r}
plot(fit2)
```

Model fit ok.

```{r}
pp_check(fit2,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "period"
)
pp_check(fit2,
  type = "dens_overlay_grouped", ndraws = 100,
  group = "trait_value"
)
```

Results:

```{r}
draw_matrix <- fit2 %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

contrast_mat <- create_contrast_matrix(fit2)

msci_draws_df2 <- as_tibble(exp(t(contrast_mat %*% t(draw_matrix))) - 1) %>%
  `colnames<-`(get_traits(fit2)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci")

summ_df2 <- msci_draws_df2 %>%
  group_by(trait) %>%
  mean_qi() %>%
  rename(mean = msci)
```

```{r}
summ_df2 %>%
  rename(msci = mean) %>%
  kable()
```

```{r}
msci_draws_df2 %>%
  full_join(summ_df2, by = join_by(trait)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  ggplot(aes(y = msci, x = trait, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(legend.position = "")
```

### Conclusion

The models perform well. We have a good fit and convergence. We can use these models to calculate the MSCI.

## Exploration Simpson's paradox

What if we compare two traits? We look at 'HostPlantSpecificity' (='Phagy') and 'NutrientValueBiotope'.

```{r}
species_traits_final %>%
  filter(trait_name %in% c("NutrientValueBiotope", "HostPlantSpecificity")) %>%
  pivot_wider(
    id_cols = c(species_nl, n_grids, period, sum_per_period),
    names_from = trait_name, values_from = trait_value
  ) %>%
  group_by(NutrientValueBiotope, HostPlantSpecificity) %>%
  mutate(n = n_distinct(species_nl)) %>%
  ungroup() %>%
  mutate(NutrientValueBiotope = factor(NutrientValueBiotope, levels = c(
    "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
    "VeryNutrientRich"
    ), ordered = TRUE)
  ) %>%
  ggplot(aes(x = NutrientValueBiotope, y = HostPlantSpecificity)) +
    geom_tile(aes(fill = n), colour = "white") +
    geom_text(aes(label = n)) +
    scale_fill_gradient(low = "blue", high = "red") +
    labs(fill = "number of\nspecies", x = "", y = "")
```

We look at 'HostPlantSpecificity' and 'NutrientValueBiotope' separately. Afterwards we combine the traits. First we look at 'HostPlantSpecificity'.

```{r}
## HostPlantSpecificity (= phagy)
dat_phagy <- species_traits_final %>%
  filter(trait_name == "HostPlantSpecificity")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit_phag <- brm(
  bf(n_grids ~ period * trait_value + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat_phagy, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit_phag",
  file_refit = "on_change"
)
```

Results:

```{r}
draw_matrix <- fit_phag %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

contrast_mat <- create_contrast_matrix(fit_phag)

msci_draws_df_phag <- as_tibble(exp(t(contrast_mat %*% t(draw_matrix))) - 1) %>%
  `colnames<-`(get_traits(fit_phag)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci")

summ_df_phag <- msci_draws_df_phag %>%
  group_by(trait) %>%
  mean_qi() %>%
  rename(mean = msci)
```

```{r}
summ_df_phag %>%
  rename(msci = mean) %>%
  kable()
```

```{r}
msci_draws_df_phag %>%
  full_join(summ_df_phag, by = join_by(trait)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  ggplot(aes(y = msci, x = trait, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(legend.position = "")
```

Then at 'NutrientValueBiotope':

```{r}
## NutrientValueBiotope
dat_ellenbergy <- species_traits_final %>%
  filter(trait_name == "NutrientValueBiotope")

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit_ellenberg <- brm(
  bf(n_grids ~ period * trait_value + offset(log(sum_per_period)) +
    (1 + period || species_nl)),
  data = dat_ellenbergy, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit_ellenberg",
  file_refit = "on_change"
)
```

Results:

```{r}
draw_matrix <- fit_ellenberg %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

contrast_mat <- create_contrast_matrix(fit_ellenberg)

msci_draws_df_ellenberg <- as_tibble(
  exp(t(contrast_mat %*% t(draw_matrix))) - 1
  ) %>%
  `colnames<-`(get_traits(fit_ellenberg)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci")

summ_df_ellenberg <- msci_draws_df_ellenberg %>%
  group_by(trait) %>%
  mean_qi() %>%
  rename(mean = msci)
```

```{r}
summ_df_ellenberg %>%
  rename(msci = mean) %>%
  kable()
```

```{r}
msci_draws_df_ellenberg %>%
  full_join(summ_df_ellenberg, by = join_by(trait)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  mutate(trait = factor(trait, levels = c(
    "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
    "VeryNutrientRich"
  ), ordered = TRUE)) %>%
  ggplot(aes(y = msci, x = trait, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(legend.position = "")
```

Do the trends change when the groups are combined (= Simpson's paradox)?

```{r}
dat3 <- species_traits_final %>%
  filter(trait_name %in% c("NutrientValueBiotope", "HostPlantSpecificity")) %>%
  pivot_wider(
    id_cols = c(species_nl, n_grids, period, sum_per_period),
    names_from = trait_name, values_from = trait_value
  )

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit3 <- brm(
  bf(n_grids ~ period * HostPlantSpecificity * NutrientValueBiotope +
       offset(log(sum_per_period)) + (1 + period || species_nl)),
  data = dat3, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit3",
  file_refit = "on_change"
)
```

Results:

```{r}
draw_matrix <- fit3 %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

contrast_mat <- rbind(
  cbind(create_contrast_matrix(3), matrix(rep(0, 9 * 3), 3)),
  matrix(cbind(
    create_contrast_matrix(3),
    matrix(c(rep(1, 3), rep(0, 6)), ncol = 3),
    contr.treatment(3),
    matrix(rep(0, 4 * 3), 3)
  ), nrow = 3),
  matrix(cbind(
    create_contrast_matrix(3),
    matrix(c(rep(0, 3), rep(1, 3), rep(0, 3)), ncol = 3),
    matrix(rep(0, 2 * 3), 3),
    contr.treatment(3),
    matrix(rep(0, 2 * 3), 3)
  ), nrow = 3),
  matrix(cbind(
    create_contrast_matrix(3),
    matrix(c(rep(0, 6), rep(1, 3)), ncol = 3),
    matrix(rep(0, 4 * 3), 3),
    contr.treatment(3)
  ), nrow = 3)
)

ellenbergn <- sort(unique(fit3$data$NutrientValueBiotope))
phagy <- sort(unique(fit3$data$HostPlantSpecificity))
col_names <- lapply(ellenbergn, function(i) {
  paste(rep(i, 3), phagy, sep = "_")
})

msci_draws_ellenberg_phagy <- as_tibble(
  exp(t(contrast_mat %*% t(draw_matrix))) - 1
  ) %>%
  `colnames<-`(unlist(col_names)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci") %>%
  separate(trait, into = c("NutrientValueBiotope", "HostPlantSpecificity")) %>%
  mutate(NutrientValueBiotope = factor(
    NutrientValueBiotope,
    levels = c(
      "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
      "VeryNutrientRich"
    ), ordered = TRUE
  ))

summ_ellenberg_phagy <- msci_draws_ellenberg_phagy %>%
  group_by(NutrientValueBiotope, HostPlantSpecificity) %>%
  mean_qi() %>%
  rename(mean = msci)
```

```{r}
summ_ellenberg_phagy %>%
  select(HostPlantSpecificity, NutrientValueBiotope, everything()) %>%
  arrange(HostPlantSpecificity, NutrientValueBiotope) %>%
  rename(msci = mean) %>%
  kable()
```

```{r}
msci_draws_ellenberg_phagy %>%
  full_join(summ_ellenberg_phagy, by = join_by(NutrientValueBiotope,
                                               HostPlantSpecificity)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  ggplot(aes(y = msci, x = NutrientValueBiotope, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(
    legend.position = "",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  facet_wrap(~HostPlantSpecificity, scales = "free")
```

### Conclusion

We see that monophagous species perform better in the second compared to the reference period. For oligophagous and polyphagous species there is no significant difference.
We see an increase in MSCI from poorer to richer nutrients.

When we combine both traits we see a more complicated picture. Monophagous species only perform better if they are linked to (very) nutrient rich habitats. Monophagous species of very nutrient poor habitats even have a significant negative MSCI. For oligophagous and polyphagous species we see no significant differences for any nutrient group. Only polyphagous species in nutrient poor habitats are less prevalent than in the reference period.

This phenomenon is called Simpson's paradox. We realize now that an increase in MSCI from poor to rich nutrients is especially pronounced for monophagous species, less for oligophagous or polyphagous species. We also observe that monophagous only have a significant higher MSCI for (very) nutrient rich habitats.

To avoid erroneous conclusions related to this paradox, we will select the $t$ most important traits that we are interested in and order these from most important to least important ($1, ..., t$). We create a model with the most important trait. Afterwards we add the second trait etc. We only keep groups containing more than 10 species.

## Exclusion of zeroes

Do we see large differences when we exclude zeroes?

```{r}
species_traits_final_nozero %>%
  filter(trait_name %in% c("NutrientValueBiotope", "HostPlantSpecificity")) %>%
  pivot_wider(
    id_cols = c(species_nl, n_grids, period, sum_per_period),
    names_from = trait_name, values_from = trait_value
  ) %>%
  group_by(NutrientValueBiotope, HostPlantSpecificity) %>%
  mutate(n = n_distinct(species_nl)) %>%
  ungroup() %>%
  mutate(NutrientValueBiotope = factor(NutrientValueBiotope, levels = c(
    "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
    "VeryNutrientRich"
    ), ordered = TRUE)
  ) %>%
  ggplot(aes(x = NutrientValueBiotope, y = HostPlantSpecificity)) +
    geom_tile(aes(fill = n), colour = "white") +
    geom_text(aes(label = n)) +
    scale_fill_gradient(low = "blue", high = "red") +
    labs(fill = "number of\nspecies", x = "", y = "")
```

We see that one group does not contain at least 10 species, but for this example we will not filter the data.

```{r}
dat_nozero <- species_traits_final_nozero %>%
  filter(trait_name %in% c("NutrientValueBiotope", "HostPlantSpecificity")) %>%
  pivot_wider(
    id_cols = c(species_nl, n_grids, period, sum_per_period),
    names_from = trait_name, values_from = trait_value
  )

# MCMC parameters
nchains <- 3 # number of chains
niter <- 10000 # number of iterations (incl. burn-in)
burnin <- niter / 4 # number of initial samples to discard (burn-in)
nparallel <- nchains # number of cores used for parallel computing
thinning <- 2

fit_nozero <- brm(
  bf(n_grids ~ period * HostPlantSpecificity * NutrientValueBiotope +
       offset(log(sum_per_period)) + (1 + period || species_nl)),
  data = dat_nozero, family = poisson(),
  chains = nchains, warmup = burnin, iter = niter, cores = nparallel,
  thin = thinning,
  file = "source/brms_cache/fit_nozero",
  file_refit = "on_change"
)
```

Results:

```{r}
draw_matrix_nozero <- fit_nozero %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

msci_draws_nozero <- as_tibble(
  exp(t(contrast_mat %*% t(draw_matrix_nozero))) - 1
  ) %>%
  `colnames<-`(unlist(col_names)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci") %>%
  separate(trait, into = c("NutrientValueBiotope", "HostPlantSpecificity")) %>%
  mutate(NutrientValueBiotope = factor(
    NutrientValueBiotope,
    levels = c(
      "VeryNutrientPoor", "NutrientPoor", "NutrientRich",
      "VeryNutrientRich"
    ), ordered = TRUE
  ))

summ_ellenberg_phagy_nozero <- msci_draws_nozero %>%
  group_by(NutrientValueBiotope, HostPlantSpecificity) %>%
  mean_qi() %>%
  rename(mean = msci)
```

```{r}
summ_ellenberg_phagy_nozero %>%
  select(HostPlantSpecificity, NutrientValueBiotope, everything()) %>%
  arrange(HostPlantSpecificity, NutrientValueBiotope) %>%
  rename(msci = mean) %>%
  kable()
```

```{r}
msci_draws_nozero %>%
  full_join(summ_ellenberg_phagy_nozero, by = join_by(NutrientValueBiotope,
                                                      HostPlantSpecificity)) %>%
  mutate(fill_var = case_when(
    .lower > 0 ~ "chartreuse3",
    .upper < 0 ~ "coral",
    .lower < 0 & .lower > -0.2 & .upper > 0 & .upper < 0.25 ~ "gold",
    TRUE ~ "skyblue",
  )) %>%
  ggplot(aes(y = msci, x = NutrientValueBiotope, fill = fill_var)) +
  stat_halfeye() +
  geom_hline(yintercept = 0, linetype = 2) +
  geom_hline(yintercept = c(-0.2, 0.25), linetype = 3) +
  scale_fill_identity() +
  labs(x = "", y = "Multispecies Change Index") +
  theme(
    legend.position = "",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  ) +
  facet_wrap(~HostPlantSpecificity, scales = "free")
```

There are differences, but they are not large.


## General conclusion

-   The models fit very well with the data in each case. This indicates that the specified Poisson model performs well.
-   Selection of a limited number of (combinations of) traits needs to be done a priori based on expert knowledge
-   Inclusion of zeroes


# Figures

We test different visualisations.
We have the following table.

```{r}
# prepare dataset
n_species_df2 <- dat2 %>%
  distinct(Activity = trait_value, n_species = n_spec)

summary_figtest <- msci_draws_df2 %>%
  group_by(trait) %>%
  median_qi(.width = 0.9) %>%
  rename(msci.median = msci,
         msci.lower = .lower,
         msci.upper = .upper,
         Activity = trait) %>%
  full_join(n_species_df2, by = join_by(Activity))

msci_draws_figtest <- msci_draws_df2 %>%
  rename(Activity = trait) %>%
  full_join(summary_figtest, by = join_by(Activity))
```

```{r}
# create dataframe to plot number of species
min_threshold <- -0.2
max_threshold <- 0.25
mult <- 0.4
min_df <- msci_draws_figtest %>%
  mutate(min = ifelse(min(msci) <= min_threshold, min(msci),
                      min_threshold),
         min = min + min * mult) %>%
  select(all_of("Activity"), n_species, min) %>%
  distinct()

msci_draws_figtest <- msci_draws_figtest %>%
  mutate(
    effects = case_when(
        msci.lower > 0 ~ "increase",
        msci.upper < 0 ~ "decrease",
        msci.lower < 0 & msci.lower > -0.2 &
          msci.upper > 0 & msci.upper < 0.25 ~ "stable",
        TRUE ~ "uncertain",
    ),
    fill_var = factor(effects,
      levels = c("decrease",
                 "increase",
                 "stable",
                 "uncertain"),
      ordered = TRUE)
  )
```

```{r}
library(INBOtheme)
theme_set(theme_inbo(base_size = 12, transparent = TRUE))
```

- half distribution and 2 intervals vertical

```{r}
msci_draws_figtest %>%
  ggplot(aes(x = Activity)) +
    stat_slab(aes(y = msci, fill = fill_var)) +
    geom_hline(yintercept = 0, linetype = "longdash", colour = "black") +
    geom_hline(yintercept = c(min_threshold, max_threshold),
               linetype = "dotdash") +
    stat_pointinterval(aes(y = msci, fill = fill_var), .width = c(0.6, 0.9)) +
    geom_label(data = min_df, aes(y = min, label = paste("n =", n_species)),
               size = 3) +
    scale_fill_manual(values =  c("firebrick1",
                                  "chartreuse3",
                                  "gold",
                                  "skyblue"),
                      drop = FALSE) +
    labs(x = "",
         y = "Multi-Species Change Index",
         fill = "") +
    theme(
      legend.position = "bottom"
    )
```

- half distribution and 2 intervals horizontal

```{r}
p_halfeye <- msci_draws_figtest %>%
  mutate(Activity = paste0(Activity, "\n", "(n = ", n_species, ")")) %>%
  ggplot(aes(y = Activity)) +
    stat_slab(aes(x = msci, fill = fill_var), orientation = "horizontal") +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
               linetype = "dotdash") +
    stat_pointinterval(aes(x = msci),
                       orientation = "horizontal", .width = c(0.6, 0.9)) +
    scale_fill_manual(values =  c("firebrick1",
                                  "chartreuse3",
                                  "gold",
                                  "skyblue"),
                      drop = FALSE) +
    labs(y = "",
         x = "Multi-Species Change Index",
         fill = "") +
    theme(
      legend.position = "bottom"
    )
p_halfeye
```

- violin and whiskers vertical

```{r}
msci_draws_figtest %>%
  ggplot(aes(x = Activity)) +
    geom_violin(aes(y = msci, fill = fill_var), colour = alpha("white", 0)) +
    geom_hline(yintercept = 0, linetype = "longdash", colour = "black") +
    geom_hline(yintercept = c(min_threshold, max_threshold),
               linetype = "dotdash") +
    geom_errorbar(data = summary_figtest, linewidth = 1, colour = "black",
                  aes(ymin = msci.lower, ymax = msci.upper), width = 0.1) +
    geom_point(data = summary_figtest, size = 2.2, colour = "black",
                  aes(y = msci.median)) +
    geom_label(data = min_df, aes(y = min, label = paste("n =", n_species)),
               size = 3) +
    scale_fill_manual(values =  c("firebrick1",
                                  "chartreuse3",
                                  "gold",
                                  "skyblue"),
                      drop = FALSE) +
    labs(x = "",
         y = "Multi-Species Change Index",
         fill = "") +
    theme(
      legend.position = "bottom"
    )
```

- violin and whiskers horizontal

```{r}
p_final <- msci_draws_figtest %>%
  mutate(Activity = paste0(Activity, "\n", "(n = ", n_species, ")")) %>%
  ggplot(aes(y = Activity)) +
    geom_violin(aes(x = msci, fill = fill_var), colour = alpha("white", 0)) +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
               linetype = "dotdash") +
    geom_errorbar(data = summary_figtest %>%
                    mutate(Activity = paste0(Activity,
                             "\n", "(n = ", n_species, ")")),
                  linewidth = 1, colour = "black",
                  aes(xmin = msci.lower, xmax = msci.upper), width = 0.1) +
    geom_point(data = summary_figtest %>%
                    mutate(Activity = paste0(Activity,
                             "\n", "(n = ", n_species, ")")),
               size = 2.2, colour = "black",
               aes(x = msci.median)) +
    scale_fill_manual(values =  c("firebrick1",
                                  "chartreuse3",
                                  "gold",
                                  "skyblue"),
                      drop = FALSE) +
    labs(y = "",
         x = "Multi-Species Change Index",
         fill = "") +
    theme(
      legend.position = "bottom"
    )
p_final
```


- add thresholds

```{r}
p_finalfinal <- p_final +
  scale_x_continuous(limits = c(-1, 1),
                     breaks = c(seq(-1, 1, 0.25)[-4], -0.2)) +
    annotate("text", label = c("increase", "strong increase"),
             x = c(0 + 0.02, max_threshold + 0.02), y = c(0.5, 0.5),
             size = 2, hjust = "left") +
    annotate("segment",
             x = c(0 + 0.02, max_threshold + 0.02), y = c(0.55, 0.55),
             xend = c(0 + 0.1, max_threshold + 0.1), yend = c(0.55, 0.55),
             arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
    annotate("text", label = c("decrease", "strong decrease"),
             x = c(0 - 0.02, min_threshold - 0.02), y = c(0.5, 0.5),
             size = 2, hjust = "right") +
    annotate("segment",
             x = c(0 - 0.02, min_threshold - 0.02), y = c(0.55, 0.55),
             xend = c(0 - 0.1, min_threshold - 0.1), yend = c(0.55, 0.55),
             arrow = arrow(type = "closed", length = unit(0.01, "npc")))
p_finalfinal
```


```{r}
p_final +
  scale_x_continuous(limits = c(-1, 1),
                     breaks = c(seq(-1, 1, 0.25)[-4], -0.2)) +
  annotate("text", y = c(0.6, 0.6), x = c(0 + 0.02, max_threshold + 0.02),
           label = c("moderate\nincrease", "strong increase"), size = 2,
           hjust = "left", vjust = "top") +
  annotate("segment", y = c(0.65, 0.65), x = c(0 + 0.02, max_threshold + 0.02),
           xend = c(0 + 0.1, max_threshold + 0.1), yend = c(0.65, 0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("text", y = c(0.6, 0.6), x = c(0 - 0.02, min_threshold - 0.02),
           label = c("moderate\ndecrease", "strong decrease"), size = 2,
           hjust = "right", vjust = "top") +
  annotate("segment", y = c(0.65, 0.65), x = c(0 - 0.02, min_threshold - 0.02),
           xend = c(0 - 0.1, min_threshold - 0.1), yend = c(0.65, 0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc")))
```

```{r}
p_final +
  scale_x_continuous(limits = c(-1, 1),
                     breaks = c(seq(-1, 1, 0.25)[-4], -0.2)) +
  annotate("text",
           y = c(0.6, 0.6), x = c(max_threshold / 2, max_threshold + 0.02),
           label = c("moderate\nincrease", "strong increase"), size = 2,
           hjust = c("center", "left"), vjust = "top") +
  annotate("segment",
           y = c(0.65, 0.65), x = c(0 + 0.02, max_threshold + 0.02),
           xend = c(0 + 0.07, max_threshold + 0.1), yend = c(0.65, 0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("segment", y = c(0.65), x = c(max_threshold - 0.02),
           xend = c(max_threshold - 0.07), yend = c(0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("text",
           y = c(0.6, 0.6), x = c(min_threshold / 2, min_threshold - 0.02),
           label = c("moderate\ndecrease", "strong decrease"), size = 2,
           hjust = c("center", "right"), vjust = "top") +
  annotate("segment",
           y = c(0.65, 0.65), x = c(0 - 0.02, min_threshold - 0.02),
           xend = c(0 - 0.07, min_threshold - 0.1), yend = c(0.65, 0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("segment", y = c(0.65), x = c(min_threshold + 0.02),
           xend = c(min_threshold + 0.07), yend = c(0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc")))
```

```{r}
p_final +
  scale_x_continuous(limits = c(-1, 1),
                     breaks = c(seq(-1, 1, 0.25)[-4], -0.2)) +
  annotate("text",
           y = c(0.6, 0.6), x = c(max_threshold / 2, max_threshold + 0.02),
           label = c("moderate\nincrease", "strong increase"), size = 2,
           hjust = c("center", "left"), vjust = "top") +
  annotate("segment", y = c(0.65), x = c(max_threshold + 0.02),
           xend = c(max_threshold + 0.1), yend = c(0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("segment", y = c(0.65), x = c(0 + 0.02),
           xend = c(max_threshold - 0.02), yend = c(0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"),
                         ends = "both")) +
  annotate("text",
           y = c(0.6, 0.6), x = c(min_threshold / 2, min_threshold - 0.02),
           label = c("moderate\ndecrease", "strong decrease"), size = 2,
           hjust = c("center", "right"), vjust = "top") +
  annotate("segment", y = c(0.65), x = c(min_threshold - 0.02),
           xend = c(min_threshold - 0.1), yend = c(0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("segment", y = c(0.65), x = c(0 - 0.02),
           xend = c(min_threshold + 0.02), yend = c(0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"),
                         ends = "both"))
```

- half distribution with annotation

```{r}
p_halfeye +
  scale_x_continuous(limits = c(-1, 1),
                     breaks = c(seq(-1, 1, 0.25)[-4], -0.2)) +
  annotate("text",
           y = c(0.6, 0.6), x = c(max_threshold / 2, max_threshold + 0.02),
           label = c("moderate\nincrease", "strong increase"), size = 2,
           hjust = c("center", "left"), vjust = "top") +
  annotate("segment", y = c(0.65), x = c(max_threshold + 0.02),
           xend = c(max_threshold + 0.1), yend = c(0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("segment", y = c(0.65), x = c(0 + 0.02),
           xend = c(max_threshold - 0.02), yend = c(0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"),
                         ends = "both")) +
  annotate("text",
           y = c(0.6, 0.6), x = c(min_threshold / 2, min_threshold - 0.02),
           label = c("moderate\ndecrease", "strong decrease"), size = 2,
           hjust = c("center", "right"), vjust = "top") +
  annotate("segment", y = c(0.65), x = c(min_threshold - 0.02),
           xend = c(min_threshold - 0.1), yend = c(0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
  annotate("segment", y = c(0.65), x = c(0 - 0.02),
           xend = c(min_threshold + 0.02), yend = c(0.65),
           arrow = arrow(type = "closed", length = unit(0.01, "npc"),
                         ends = "both"))
```


## Conclusion

We choose this layout as final:

```{r}
p_finalfinal
```

## Log-scale

Can we make the figures symmetrical in the log-scale?

```{r}
draw_matrix <- fit2 %>%
  spread_draws(`b_periodp2013_2022.*`, regex = TRUE) %>%
  select(starts_with("b_periodp2013_2022")) %>%
  as.matrix()

contrast_mat <- create_contrast_matrix(fit2)

msci_draws_log <- as_tibble(t(contrast_mat %*% t(draw_matrix))) %>%
  `colnames<-`(get_traits(fit2)) %>%
  pivot_longer(cols = everything(), names_to = "trait", values_to = "msci")

summary_figtest2 <- msci_draws_log %>%
  group_by(trait) %>%
  median_qi(.width = 0.9) %>%
  rename(msci.median = msci,
         msci.lower = .lower,
         msci.upper = .upper,
         Activity = trait) %>%
  full_join(n_species_df2, by = join_by(Activity))

msci_draws_figtest2 <- msci_draws_log %>%
  rename(Activity = trait) %>%
  full_join(summary_figtest2, by = join_by(Activity))
```

```{r}
# create dataframe to plot number of species
min_threshold <- log(0.8)
max_threshold <- log(1.25)
mult <- 0.4
min_df2 <- msci_draws_figtest2 %>%
  mutate(min = ifelse(min(msci) <= min_threshold, min(msci),
                      min_threshold),
         min = min + min * mult) %>%
  select(all_of("Activity"), n_species, min) %>%
  distinct()

msci_draws_figtest2 <- msci_draws_figtest2 %>%
  mutate(
    effects = case_when(
        msci.lower > 0 ~ "increase",
        msci.upper < 0 ~ "decrease",
        msci.lower < 0 & msci.lower > min_threshold &
          msci.upper > 0 & msci.upper < max_threshold ~ "stable",
        TRUE ~ "uncertain",
    ),
    fill_var = factor(effects,
      levels = c("increase",
                 "stable",
                 "decrease",
                 "uncertain"),
      ordered = TRUE)
  )
```

https://github.com/inbo/abvanalysis/blob/1540393796b92f96833f596e331d44de1363af79/inst/website/_soort.Rmd

```{r}
index_labels <- function(x) {
  sprintf("%+.0f%%", 100 * (exp(x) - 1))
}


index_breaks <- function(x) {
  z <- 1 - c(
    9 / 10, 4 / 5, 3 / 4, 2 / 3, 1 / 2, 1 / 3, 1 / 4, 1 / 5, 1 / 10, 0
  )
  z <- log(sort(z))
  z <- z[which(z >= min(-abs(x)))[1] + -1:3]
  c(z, 0, -z)
}

minor_breaks <- function(x) {
  breaks <- unique(abs(index_breaks(x)))
  out <- vector(length = length(breaks) - 1)
  i <- 1
  while (i < length(breaks)) {
    out[i] <- mean(c(breaks[i], breaks[i + 1]))
    i <- i + 1
  }
  return(c(-out, out))
}

index_breaks_rev <- function(x) {
  exp(index_breaks(x))
}

index_labels_rev <- function(x, digits = 2) {
  format(round(exp(index_breaks(x)), digits), nsmall = digits)
}
```

```{r}
msci_draws_figtest2 %>%
  mutate(Activity = paste0(Activity, "\n", "(n = ", n_species, ")")) %>%
  ggplot(aes(y = Activity)) +
    geom_violin(aes(x = msci, fill = fill_var), colour = alpha("white", 0)) +
    geom_vline(xintercept = 0, linetype = "longdash", colour = "black") +
    geom_vline(xintercept = c(min_threshold, max_threshold),
               linetype = "dotdash") +
    geom_errorbar(data = summary_figtest2 %>%
                    mutate(Activity = paste0(Activity,
                             "\n", "(n = ", n_species, ")")),
                  linewidth = 1, colour = "black",
                  aes(xmin = msci.lower, xmax = msci.upper), width = 0.1) +
    geom_point(data = summary_figtest2 %>%
                    mutate(Activity = paste0(Activity,
                             "\n", "(n = ", n_species, ")")),
               size = 2.2, colour = "black",
               aes(x = msci.median)) +
    scale_fill_manual(values =  c("chartreuse3",
                                  "gold",
                                  "firebrick1",
                                  "skyblue"),
                      drop = FALSE) +
    labs(y = "",
         x = "Multi-Species Change Index",
         fill = "Trend:") +
    theme(
      legend.position = "bottom"
    ) +
    annotate("text", label = c("increase", "strong increase"),
             x = c(0 + 0.02, max_threshold + 0.02), y = c(0.5, 0.5),
             size = 2, hjust = "left") +
    annotate("segment",
             x = c(0 + 0.02, max_threshold + 0.02), y = c(0.55, 0.55),
             xend = c(0 + 0.1, max_threshold + 0.1), yend = c(0.55, 0.55),
             arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
    annotate("text", label = c("decrease", "strong decrease"),
             x = c(0 - 0.02, min_threshold - 0.02), y = c(0.5, 0.5),
             size = 2, hjust = "right") +
    annotate("segment",
             x = c(0 - 0.02, min_threshold - 0.02), y = c(0.55, 0.55),
             xend = c(0 - 0.1, min_threshold - 0.1), yend = c(0.55, 0.55),
             arrow = arrow(type = "closed", length = unit(0.01, "npc"))) +
    scale_x_continuous(breaks = index_breaks, labels = index_labels,
                       minor_breaks = minor_breaks,
                       limits = c(-log(2), log(2)),
                       sec.axis = sec_axis(~ exp(.),
                        breaks = index_breaks_rev(msci_draws_figtest2$msci),
                        labels = index_labels_rev(msci_draws_figtest2$msci),
                        name = "Proportional change"))
```
